---
title: "Data Analysis"
author: "Simon Gendrisch"
format: html
editor: source
---

```{r, packages}
library(tidyverse)
library(sf)
library(dplyr)
library(Distance)
library(mrds)
source("./utils.R")
```

```{r, Load layers}

# QField points
qfield_wp <- st_read("./data/GIS/test/survey_data.gpkg", layer = "survey_data")

# Load test data
test <- read.csv("./data/GIS/test/observations_filled.csv")

```

Calculate distance from reticles or angle (in degree) and observer height (in m).

```{r}

obs_height = 2    # Observer height in meters

qfield_obs <- qfield_wp %>%  
  mutate(
    distance = case_when(
      !is.na(reticles) ~ calculate_distance_bino(reticles = reticles, height = obs_height), # If reticles available reticles will be used to calculate distance
      !is.na(vert_angle) ~ calculate_distance_inclino(angle = vert_angle, height = obs_height),
      TRUE ~ NA_real_
    ) ) %>% 
   mutate(
    method_used = case_when(
      !is.na(reticles) ~ "reticle", 
      !is.na(vert_angle) ~ "angle",
      TRUE ~ "missing"
    )
  )

#### TEST #####
test_dis <- test %>%  
  mutate(
    distance = case_when(
      !is.na(reticles) ~ calculate_distance_bino(reticles = reticles, height = obs_height), # If reticles available reticles will be used to calculate distance
      !is.na(vert_angle) ~ calculate_distance_inclino(angle = vert_angle, height = obs_height),
      TRUE ~ NA_real_
    ) ) %>% 
   mutate(
    method_used = case_when(
      !is.na(reticles) ~ "reticle", 
      !is.na(vert_angle) ~ "angle",
      TRUE ~ "missing"
    )
  )

# Change bird and fish presence to binary
test_dis <- test_dis %>% 
  mutate(bird_presence = case_when(
    bird_presence == "False" ~ 0,
    bird_presence == "True" ~ 1,
    TRUE ~ NA_real_
  )) %>% 
  mutate(fish_presence = case_when(
    fish_presence == "False" ~ 0,
    fish_presence == "True" ~ 1,
    TRUE ~ NA_real_
  ))
 

```

Plot data as Histogram to see possible observer biases.

```{r}
# Hist plot to see potential biases; should start high and drop after a broad shoulder
test_dis %>% 
  ggplot() +
  geom_histogram(bins = 5, aes(     # Adjust bins to data
    x = distance))
  
# Boxplot to identify biases for specific species (bigger species easier to detect)
test_dis %>% 
  ggplot(aes(
    x = sort(distance),
    y = species,
    fill = species
  )) +
  geom_boxplot(alpha = .5)

```

Using the `Distance` package to run different models and see which one works best. **Low `AIC` is better**. `Distance` needs ceratain information and dataframe format to run the model: Region.Label \| Area \| Sample.Label \| Effort \| object \| distance \| Study.Area.

Drop Observer Column for models.

```{r, Run Distance Models}

test_new <-test_dis %>% 
  select(!observer)


model_hn <- ds(
  data = test_new,
  truncation = 30,    # Adjusted depending on data
  key = "hn"          # Use differenet models to find best fitting one
)


summary(model_hn)
plot(model_hn)

  model_hr <- ds(
  data = test_new,
  truncation = 30,
  key = "hr"
)

summary(model_hr)
plot(model_hr)


AIC(model_hn, model_hr)    # Use to compare different models

```



