---
title: "Distance Sampling"
author: "Simon Gendrisch"
date: "2026-02-16"
output: html_document
---

```{r packages}
library(dssd)
library(dsims)
library(here)
library(tidyverse)
library(sf)
library(mrds)
library(Distance)
library(nlme)
library(mgcv)

```

```{r, loading shapefile}
shapefile <- st_read("data/GTW.shp")

shape_split <- st_read("data/GTW_split.gpkg")

# Transform degree in meter
shapefile_tr <- st_transform(shapefile, 32630)
shape_split_tr <- st_transform(shape_split, 32630)

# Test if in right format
st_bbox(shapefile_tr)
st_bbox(shape_split_tr)


region <- make.region(region.name = "Gibraltar",
                      shape = shapefile_tr,
                      units = "m")

region_split <- make.region(region.name = "Gibraltar",
                      shape = shape_split_tr,
                      units = "m")


plot(region)
plot(region_split)


```

```{r, Create Density}

# We first create a flat density grid
density <- make.density(region = region,
                        x.space = 500,
                        constant = 1)

# Now we can add some high and low points to give some spatial variability
density <- add.hotspot(object = density,
                       centre = c(286000, 3998000),
                       sigma = 8000,
                       amplitude = 4)

density <- add.hotspot(object = density,
                       centre = c(292000, 4000000),
                       sigma = 10000,
                       amplitude = -1)


plot(density, region)

density.data <- density@density.surface[[1]]

fit.gam <- gam(density ~ s(x,y), data = density.data, family = gaussian(link="log"))

# Use the gam object to create a density object
gam.density <- make.density(region = region,
                            x.space = 500,
                            fitted.model = fit.gam)

plot(gam.density, region, scale = 0.001)

```

```{r, Defining a Population}

# Create a covariate list describing the distribution of cluster sizes
covariates <- list(size = list(distribution = "ztruncpois", mean = 3))

# Define the population description
pop.desc <- make.population.description(region = region,
                                        density = gam.density,
                                        covariates = covariates,
                                        N = 300,
                                        fixed.N = TRUE)


```

```{r, Define Design}


parallel.design <- make.design(region = region, 
                               design = "systematic",
                               spacing = 500,
                               edge.protocol = "minus",
                               design.angle = 45,
                               truncation = 750)

p.survey <- generate.transects(parallel.design)
plot(region, p.survey)



zigzag.design <- make.design(region = region, 
                             design = "eszigzag",
                             spacing = 500,
                             edge.protocol = "minus",
                             design.angle = 0,
                             bounding.shape = "convex.hull",
                             truncation = 750)

z.survey <- generate.transects(zigzag.design)
plot(region, z.survey)

```

# Defining Detectability

Once we have defined both the population of interest and the design which we will use to survey our population we now need to provide information about how detectable the individuals or clusters are. For this example we will assume that larger clusters are more detectable. Take care when defining covariate parameters that the covariate names match those in the population description

```{r, Covariate}

# Define the covariate parameters on the log scale
cov.param <- list(size = log(1.08))

# Create the detectability description
detect <- make.detectability(key.function = "hn",
                             scale.param = 300,
                             cov.param = cov.param,
                             truncation = 750)

# Plot the simulation detection functions
plot(detect, pop.desc)


```

# Defining Analyses

The final component to a simulation is the analysis or set of analyses you wish to fit to the simulated data. We will define a number of models and allow automatic model selection based on the minimum AIC value. The models included below are a half-normal with no covariates, a hazard rate with no covariates and a half-normal with cluster size as a covariate. We will leave the truncation value at 750 as previously defined (it must be $\le$ to the truncation values used previously). We will use the default error variance estimator “R2”. See ?mrds::varn for descriptions of the various empirical variance estimators for encounter rate.

```{r, Analysis}

analyses <- make.ds.analysis(dfmodel = list(~1, ~1, ~size),
                             key = c("hn", "hr", "hn"),
                             truncation = 750,
                             er.var = "R2",
                             criteria = "AIC")

```

```{r, Simulation}

sim.parallel <- make.simulation(reps = 99,
                                design = parallel.design,
                                population.description = pop.desc,
                                detectability = detect,
                                ds.analysis = analyses)

sim.zigzag <- make.simulation(reps = 99,
                              design = zigzag.design,
                              population.description = pop.desc,
                              detectability = detect,
                              ds.analysis = analyses)


# Generate a single instance of a survey: a population, set of transects 
# and the resulting distance data
eg.parallel.survey <- run.survey(sim.parallel)

# Plot it to view a summary
plot(eg.parallel.survey, region)


# Generate a single instance of a survey: a population, set of transects 
# and the resulting distance data
eg.zigzag.survey <- run.survey(sim.zigzag)

# Plot it to view a summary
plot(eg.zigzag.survey, region)

```

# Running the Simulations

```{r, Simulations}

# Running the simulations
sim.parallel <- run.simulation(sim.parallel)
summary(sim.parallel)

sim.zigzag <- run.simulation(sim.zigzag)
summary(sim.zigzag)


```

# Using dssd package

```{r, Workflow with dssd}

library(dssd)
library(sf)

plot(region)

cover <- make.coverage(region,n.grid.points = 1000)

plot(region, cover)

design <- make.design(
  region = region,
  line.length = 0.002,
  design.angle =  90,
  edge.protocol = "minus",
  truncation = 1,
  design = "systematic",
  coverage.grid = cover
)

transects <- generate.transects(design)
transects

plot(region, transects)

design <- run.coverage(design, reps = 10)

st_bbox(shapefile)

```

# Run Claude version
Instead of Chat, use Claude AI to give me better results

```{r, Claude example}

library(dssd)
library(sf)

# Import shape file and convert to region

shapefile <- st_read("data/GTW.shp")

# Transform degree in meter
shapefile_tr <- st_transform(shapefile, crs = 32630)

# Test if in right format
st_bbox(shapefile_tr)


region <- make.region(region.name = "Gibraltar",
                      shape = shapefile_tr,
                      units = "m")

plot(region)



```

```{r, Study Design}

study_design <- make.design(
  region = region_split,
  transect.type =  "line",
  design = "eszigzag",
  spacing = c(1000, 1000, 1000),
  design.angle = c(90, 0, 0),
  edge.protocol = "minus",
  truncation = 500
)

transects_c <- generate.transects(study_design)

plot(region_split, transects_c)

coverage <- run.coverage(
              design = study_design,
              reps = 10
)

plot(coverage)
summary(coverage)




```



```{r}
# Test a range of spacings and compare coverage
spacings <- c(750, 500, 350)
results <- data.frame(spacing = spacings, mean_cov = NA, cv = NA, total_effort_km = NA)

for (i in seq_along(spacings)) {
  d <- make.design(
    region        = region_split,
    transect.type = "line",
    design        = "eszigzag",
    spacing       = spacings[i],
    design.angle  = c(90, 0, 0),
    edge.protocol = "minus",
    truncation    = 300
  )
  
  cov <- run.coverage(d, reps = 100)
  t   <- generate.transects(d)
  
  scores <- cov@coverage.scores
  results$mean_cov[i]       <- round(mean(scores), 3)
  results$cv[i]             <- round(sd(scores) / mean(scores), 3)
  results$total_effort_km[i] <- round(sum(t@samplers$length) / 1000, 1)
}

print(results)
```

  spacing mean_cov    cv total_effort_km
1    1000    0.966 0.144               0
2    1500    0.654 0.146               0
3    2000    0.506 0.148               0
4    2500    0.416 0.161               0
5    3000    0.357 0.195               0

