---
title: "Results"
format: docx
execute:
  echo: false              
  warning: false
  message: false
---

```{r, Packages}
library(tidyverse)
library(readxl)
library(lubridate)
library(FSA)
library(flextable)
```

```{r, Read data}
#| include: false
# Excel sheet of the spots with details
spots <- read_excel(path = "./data/datasheet_landsurvey.xlsx", sheet = "spots")


# Clean the sighting data and import
source("./land_survey.R")

full <- readRDS("./temp/full.rds")


# Set active spots as variable
active_spots <- spots %>%  
  filter(spot_name != "MedSteps" & spot_name != "WestSide") %>% 
  select(spot_name)

odontoceti <- c("common_dolphin", "striped_dolphin", "bottlenose_dolphin", "delphinidae")
mysteciti <- c("fin_whale")

```

```{r}
#|include: false

by_species <- data %>% 
  summarise(
    n_species = n(),
    .by = "species"
  ) %>% 
  pivot_wider(
    names_from = species,
    values_from = n_species
  )

by_familiy <- full %>% 
  filter(source == "land") %>% 
  mutate(family = case_when(
    species %in% odontoceti ~ "odontoceti",
    species %in% mysteciti ~ "mysteciti",
    TRUE ~ NA_character_)) %>% 
  summarise(
    sighting = n(),
    .by = "family"
  ) %>% 
  pivot_wider(
    names_from = family,
    values_from = sighting
  )


per_spot <- data %>% 
  summarise(
    sightings = n(),
    .by = "spot_name"
  )


full_source <- full %>% 
  summarise(
    sightings = n(),
    .by = "source"
  ) %>% 
  pivot_wider(
    names_from = source,
    values_from = sightings
  )

full_sum <- full %>% 
  summarise(
    sightings = n()
  )

boat_species <- full %>% 
  filter(source == "boat") %>% 
  summarise(
    n_species = n(),
    .by = "species"
  ) %>% 
  pivot_wider(
    names_from = species,
    values_from = n_species
  )

boat <- full %>% 
  filter(source == "boat")


```

```{r, Effort analysis}
#|include: false

effort <- effort_log %>% 
  mutate(across(c(observation_start, observation_end),
                ~ hms::as_hms(.))) %>% 
  filter(!grepl("off_effort", survey_id)) %>% 
  mutate(effort_time = as.duration(observation_end - observation_start))

er <- data %>%  
  filter(spot_name %in% active_spots$spot_name) %>% 
  summarise(max_group = max(group_size, na.rm = TRUE),
            min_group = min(group_size, na.rm = TRUE),
            max_dis = max(distance, na.rm = TRUE),
            min_dis = min(distance, na.rm = TRUE),
            sightings = n(),
            .by = "spot_name") %>%
  left_join(effort %>%
              group_by(spot_name) %>% 
  summarise(hours = sum(effort_time)/60^2), by = "spot_name") %>% 
  mutate(per_hour = round(sightings/hours, 2)) %>% 
  select(spot_name, sightings, hours, per_hour, max_dis, min_dis, max_group, min_group) %>% 
  arrange(desc(per_hour))


total_land <- effort %>% 
  summarise(
    surveys = n()
  )




```

```{r}
#| include: false
#| output: false

effort <- effort_log %>%
  filter(!str_detect(survey_id, "off_effort")) %>%
  mutate(
    duration_min = as.numeric(difftime(observation_end, observation_start,
                                        units = "mins"))
  )
 
# ---- 3. Detections: keep on-effort sightings -------------------------------
sightings <- detection %>%
  filter(on_effort == 1)

 
# ---- 4. Sightings & individuals per scan -----------------------------------
per_scan <- sightings %>%
  group_by(survey_id) %>%
  summarise(
    n_sightings   = n(),
    n_individuals = sum(group_size, na.rm = TRUE),  # one NA group size -> 0
    .groups = "drop"
  )
 
# Join back onto the full effort table; scans with no sighting become 0 (NOT NA)
effort <- effort %>%
  left_join(per_scan, by = "survey_id") %>%
  mutate(
    n_sightings   = replace_na(n_sightings, 0),
    n_individuals = replace_na(n_individuals, 0)
  )
 
# ---- 5. SPUE by spot -------------------------------------------------------
spue_by_spot <- effort %>%
  group_by(spot_name) %>%
  summarise(
    scans        = n(),
    effort_hours = sum(duration_min, na.rm = TRUE) / 60,
    sightings    = sum(n_sightings),
    individuals  = sum(n_individuals),
    .groups = "drop"
  ) %>%
  mutate(
    spue_per_scan   = sightings / scans,        # headline SPUE
    sightings_per_h = sightings / effort_hours, # encounter rate
    indiv_per_h     = individuals / effort_hours
  ) %>%
  arrange(desc(scans))
 
# ---- 6. SPUE overall -------------------------------------------------------
spue_overall <- effort %>%
  summarise(
    spot_name    = "TOTAL",
    scans        = n(),
    effort_hours = sum(duration_min, na.rm = TRUE) / 60,
    sightings    = sum(n_sightings),
    individuals  = sum(n_individuals)
  ) %>%
  mutate(
    spue_per_scan   = sightings / scans,
    sightings_per_h = sightings / effort_hours,
    indiv_per_h     = individuals / effort_hours
  )
 
spue_summary <- bind_rows(spue_by_spot, spue_overall)
 
# ---- 7. Report & export ----------------------------------------------------
spue_report <- spue_summary %>%
  mutate(across(c(effort_hours, spue_per_scan, sightings_per_h, indiv_per_h),
                ~ round(.x, 3)))
 
print(spue_report, n = Inf)



```

```{r, Kruskal-Wallis and Dunn test}
 #|include: false

# Include the right df from datavis script

 active <- c("Europa", "SandyBay", "PrinceGeorge")
 scans  <- effort %>% filter(spot_name %in% active)   # per-scan rows, NOT the summary

# count(scans, spot_name)


 SPUE_kru <- kruskal.test(n_sightings ~ spot_name, data = scans)

 SPUE_dunn <- dunnTest(n_sightings ~ factor(spot_name), data = scans, method = "bh")

```

This study adapted point transect distance sampling to small cetaceans, utilising a modified version of the point transect protocol primarily employed for birds or larger cetaceans (**REFS**). The protocol was adapted to incorporate land-based surveys, enabling a systematic assessment of the cetacean distribution in the BGTW.

Overall, `r  full_sum$sightings` cetacean sightings were recorded between `r format(min(full$date.x), "%d.%m.%Y")` and `r format(max(full$date.x), "%d.%m.%Y")`, including pilot studies and both vessel- and land-based surveys. A total of 2 vessel-based surveys was conducted, accounting for `r full_source$boat` of these sightings: `r boat_species$common_dolphin` common dolphins (*Delphinus delphis*) and `r boat_species$striped_dolphin` striped dolphins (*Stenella coeruleoalba*) and `r boat_species$delphinidae` sighting where the species could not be identified. The group size for vessel-based sightings ranged from `r min(boat$group_size, na.rm = TRUE)` to `r max(boat$group_size, na.rm = TRUE)` (mean = `r round(mean(boat$group_size, na.rm = TRUE), 2)`, ±`r round(sd(boat$group_size, na.rm = TRUE), 2)`, median = `r round(median(boat$group_size, na.rm = TRUE), 2)`).

In total `r total_land$surveys` land-based surveys were conducted, resulting in `r round(sum(er$hours), 2)` total on effort hours over `r length(unique(effort$date))` survey days. `r length(data$survey_id)` cetacean sightings were recorded in this sampling period, of which `r sum(data$on_effort)` were on-effort and `r sum(!data$on_effort)` off-effort. Of these, `r by_familiy$odontoceti` were Odontoceti, so either common dolphins (n = `r by_species$common_dolphin`) or dolphins where the species couldn’t certainly be defined (n = `r by_species$delphinidae`) and `r by_familiy$mysteciti` Mysticeti of which `r by_species$fin_whale` could be identified as Fin whales (*Balaenoptera physalus*). `r by_species$unidentified` sightings could not be reliably assigned and were recorded as unidentified. The group size for land-based sightings ranged from `r min(data$group_size, na.rm = TRUE)` to `r max(data$group_size, na.rm = TRUE)` (mean = `r round(mean(data$group_size, na.rm = TRUE), 2)`, ±`r round(sd(data$group_size, na.rm = TRUE), 2)`, median = `r round(median(data$group_size, na.rm = TRUE), 2)`). The overview table (see @tbl-spot) showed different maximum and minimum distances as well as varying group sizes across the spots.

```{r}
#| label: tbl-spot
#| tbl-cap: "Statistics per spots over the sampling period."

flextable(er) %>%
   set_header_labels(
    per_hour = "Sightings per hour",
    spot_name = "Spotname",
    sightings = "# of sightings",
    hours = "on-effort Hours",
    max_dis = "Maximum Distance",
    min_dis = "Minimum Distance",
    max_group = "Maximum Group Size",
    min_group = "Minimum Group Size"
  ) %>%
  colformat_double(
    j = "hours",
    digits = 2
  ) %>%
  colformat_double(
    j = c("max_dis", "min_dis"),
    digits = 0,
    suffix = " m"
  ) %>% 
  autofit() %>%
  theme_booktabs() %>% 
  bold(bold = TRUE, part = "header")
```

Most of the sightings were recorded at `r per_spot$spot_name[which.max(per_spot$sightings)]`, covering mainly the X side (n = `r max(per_spot$sightings)`). And between the hours XX and XX (Figure daytime sihgtings). Sightings per unit effort (SPUE) were highest at spot XX (Figure XX, Table @tbl-spue_report). A Kruskal-Wallis test showed a significant difference in SPUE among spots (χ² = `r round(as.numeric(SPUE_kru$statistic),2)`, df = `r as.numeric(SPUE_kru$parameter)`, p = `r round(SPUE_kru$p.value,3)`), Dunn’s test, with a Benjamini-Hochberg correction, showed a significant difference between spot XX and XX (adjusted p = `r SPUE_dunn$res[1,4]`). Using the Distance package (**REFS**) a detection function was fitted per spot (FigurXX). Using the detection functions the relative abundance of dolphins was calculated.

```{r}
#| label: tbl-spue_report
#| tbl-cap: "Sightings per Unit effort were different over the three spots as well as number of individuals."

spue_report %>% 
  filter(spot_name %in% active_spots$spot_name | spot_name == "TOTAL") %>% 
  select(!geometry) %>% 
  flextable() %>% 
  delete_columns(j = "geometry") %>% 
  set_header_labels(
    spot_name = "Spot Names",
    scans = "Scans",
    effort_hours = "Hours",
    sightings = "Sightings",
    individuals = "Individuals",
    spue_per_scan = "SPUE (per scan)",
    sightings_per_h = "Sightings per h",
    indiv_per_h = "Indiv. per h"
  ) %>% 
  colformat_double(
    j = c("effort_hours", "sightings_per_h", "sightings_per_h"),
    digits = 2
  ) %>% 
  bold(i = 4) %>% 
  autofit() %>% 
  theme_booktabs()
```

The Kernel Density Estimation using group size as a weight variable showed the highest density around the south western tip of the peninsula (lat: XX, lon: XX). With the core area covering most of the southern sector (Figure XX). The home area stretched around the whole peninsula covering most of the BGTW and every sector.
