---
title: "Survey Design"
author: "Simon Gendrisch"
date: "2026-03-05"
output: html_document
---

```{r, Load packages}

library(mrds)
library(Distance)
library(dssd)
library(tidyverse)
library(sf)
library(dplyr)
library(ggplot2)

source("./utils.R")



```




```{r, importing data}

shape_split <- st_read("data/GTW_split_acc.gpkg")

# Transform degree in meter
split_tr <- st_transform(shape_split, crs = 32630)

split_tr


region_split <- make.region(region.name = "Gibraltar",
                      shape = split_tr,
                      strata.name = c("South", "West", "East"),
                      units = "m")

plot(region_split)


```

The rule is strict: spacing must always be ≥ 2 × truncation to avoid overlap. For a zigzag it should be even larger because the angled legs cover more ground than straight parallel lines.


```{r, Study Design}

study_design <- make.design(
  region = region_split,
  transect.type =  "line",
  design = "eszigzag",
  spacing = c(1500, 1500, 1500),
  design.angle = c(90, 0, 0),
  edge.protocol = "minus",
  truncation = 300
)

transects_c <- generate.transects(study_design)

transects_c
plot(region_split, transects_c, covered.area = TRUE)

coverage <- run.coverage(
              design = study_design,
              reps = 15
)

plot(coverage)



summary(coverage@coverage.scores)
summary(transects_c@line.length)

summary(transects_c@samplers)




### Using seeds to create "random" transects each day -> make a list of days and asign seed to every day -> generate transects


```

# Export transect line

Export the above design to test it on GPS device. After this I will assign different seeds per day, if I want the transects for the ceratin day, including the seed I need to go into the loop and pull from there.

```{r, Export transect line for this design}


transect_export <- st_transform(transects_c@samplers, crs = 4326)

transect_export_clean <- st_geometry(transect_export)


st_write(transect_export_clean, "transects.gpx", layer = "tracks", driver = "GPX")



export_transects_gpx(transects_c, "transects_garmin.gpx")

# With a date/seed prefix for daily runs:
# export_transects_gpx(transects_c,
#                      path        = "transects_2026-04-21.gpx",
#                      name_prefix = "2026-04-21")

```


Test for feasible Spacing for the area


```{r, Spacing testing}

spacings = c(550, 600, 650, 700)      # Spacing of 667 m is ideal for the area
boat_speed = 10
results = data.frame(spacing = spacings, mean_cov=NA, cv=NA, effort_km=NA, n_transects=NA, h_survey=NA)

for (i in seq_along(spacings)){
  
  d <- make.design(
  region = region_split,
  transect.type =  "line",
  design = "eszigzag",
  spacing = spacings[i],
  design.angle = c(90, 0, 0),
  edge.protocol = "minus",
  truncation = 300
)
  
  t <- generate.transects(d)
  effort <- sum(as.numeric(sf::st_length(t@samplers))) / 1000
  
  if (effort > 0) {
    cov    <- run.coverage(d, reps=100)
    scores <- unlist(cov@coverage.scores)  # unlist across all strata
    
    results$mean_cov[i]    <- round(mean(scores), 3)
    results$cv[i]          <- round(sd(scores)/mean(scores), 3)
    results$effort_km[i]   <- round(effort, 1)
    results$n_transects[i] <- nrow(t@samplers)
    results$h_survey[i]    <- round(sum(effort/(boat_speed*1.852)), 1)
    
  } else {
    cat("Spacing", spacings[i], "— no transects generated, too large for region\n")
  }
  
}

print(results)


plot(cov)



```

  spacing mean_cov    cv effort_km n_transects h_survey
1     550    1.067 0.106     151.1          37      8.2
2     600    0.978 0.118     139.5          34      7.5
3     650    0.902 0.119     128.0          33      6.9
4     700    0.840 0.122     115.9          30      6.3




```{r}
library(dssd)
library(sf)

# ── Step 1: Rebuild design at chosen spacing ──────────────────────────────────
study_design <- make.design(
  region        = region_split,
  transect.type = "line",
  design        = "eszigzag",
  spacing       = c(650, 650, 650),
  design.angle  = c(90, 0, 0),
  edge.protocol = "minus",
  truncation    = 300
)

trans <- generate.transects(study_design)

plot(region_split, trans)

cov <- run.coverage(study_design, reps = 10)
plot(cov)

# ── Step 2: Generate a large pool of candidate realisations ───────────────────
# We'll screen these and pick the most spatially complementary ones
set.seed(42)
n_candidates  <- 200    # generate 200 candidates, pick best N from them
n_survey_days <- 15     # how many field days you're planning

candidate_seeds      <- sample(1:99999, n_candidates)
candidate_transects  <- list()
candidate_centroids  <- list()   # store mean Y position of transects per realisation

cat("Generating", n_candidates, "candidate realisations...\n")

for (i in seq_along(candidate_seeds)) {
  set.seed(candidate_seeds[i])
  t <- generate.transects(study_design)
  
  # Check it actually generated transects
  effort <- sum(as.numeric(sf::st_length(t@samplers))) / 1000
  
  if (effort > 0) {
    candidate_transects[[i]] <- t
    
    # Extract centroid of all transect lines — proxy for spatial position
    centroids <- sf::st_centroid(sf::st_geometry(t@samplers))
    coords    <- sf::st_coordinates(centroids)
    candidate_centroids[[i]] <- c(mean(coords[,1]), mean(coords[,2]))
  }
}

cat("Valid realisations:", sum(sapply(candidate_transects, \(x) !is.null(x))), "\n")

# ── Step 3: Score realisations for spatial spread ─────────────────────────────
# Extract X centroid (east-west position) for each realisation
# In a zigzag design, the random start shifts the whole pattern laterally
# We want to select seeds that are maximally spread across this axis

valid_idx <- which(!sapply(candidate_transects, is.null))
x_positions <- sapply(valid_idx, \(i) candidate_centroids[[i]][1])
y_positions <- sapply(valid_idx, \(i) candidate_centroids[[i]][2])

# Normalise positions to 0-1 range
x_norm <- (x_positions - min(x_positions)) / diff(range(x_positions))

# ── Step 4: Greedy selection — pick seeds that are maximally spread ────────────
# Start with the median position, then iteratively pick the most distant remaining

selected_idx <- integer(n_survey_days)
remaining    <- seq_along(valid_idx)

# First pick: closest to median spatial position (a "neutral" start)
selected_idx[1] <- remaining[which.min(abs(x_norm[remaining] - 0.5))]
remaining <- remaining[-selected_idx[1]]

# Subsequent picks: always choose realisation furthest from already-selected ones
for (k in 2:n_survey_days) {
  # Distance from each remaining candidate to all already-selected
  dists <- sapply(remaining, \(j) {
    min(abs(x_norm[j] - x_norm[selected_idx[1:(k-1)]]))
  })
  best  <- remaining[which.max(dists)]
  selected_idx[k] <- best
  remaining <- remaining[-which(remaining == best)]
}

# ── Step 5: Build your survey schedule ────────────────────────────────────────
selected_seeds <- candidate_seeds[valid_idx[selected_idx]]

survey_schedule <- data.frame(
  day         = 1:n_survey_days,
  seed        = selected_seeds,
  date        = NA,                # fill in as you go
  completed   = FALSE,
  effort_km   = NA,
  n_sightings = NA,
  notes       = NA
)

cat("\n=== Survey Schedule ===\n")
print(survey_schedule[, c("day", "seed")])



region_geom <- sf::st_geometry(region_split@region)

# ── Build a single long dataframe of all selected transects ──────────────────
transect_df <- lapply(1:n_survey_days, function(k) {
  i <- valid_idx[selected_idx[k]]
  t <- candidate_transects[[i]]
  
  geom <- sf::st_geometry(t@samplers)
  df   <- sf::st_sf(geometry = geom)
  df$day   <- paste0("Day ", sprintf("%02d", k))
  df$seed  <- selected_seeds[k]
  df
}) |> bind_rows()

# ── Region background as df ───────────────────────────────────────────────────
region_df <- sf::st_sf(geometry = region_geom)

# ── Facet plot ────────────────────────────────────────────────────────────────
ggplot() +
  geom_sf(data  = region_df,
          colour = "#1e3a5f",
          linewidth = 0.4) +
  geom_sf(data      = transect_df,
          colour    = transect_df$seed,
          linewidth = 0.5,
          alpha     = 0.9) +
#  facet_wrap(~day, ncol = 5) +
  labs(
    title    = "Survey Transects per Field Day",
    subtitle = paste0("Spacing = 650m | Truncation = 300m | Seed-optimised for spatial spread"),
    caption  = paste0("Seeds: ", paste(selected_seeds, collapse = ", "))
  ) +
  theme_minimal(base_size = 10) 




```



## What This Does

The key idea is that in a zigzag design, each random start **shifts the entire transect grid laterally** by a small amount. By selecting seeds whose shifts are maximally spread across the full range of possible positions, consecutive survey days fill in different "gaps" in the coverage grid:

Day 1 seed:  transects at positions  0, 650, 1300, 1950...
Day 2 seed:  transects at positions  325, 975, 1625, 2275...
Day 3 seed:  transects at positions  160, 810, 1460, 2110...
→ Together they tile the region much more evenly than random seeds would



## Optimized Survey transects

Here I optimized each transect per area instead of overall coverage to achieve better scores.


```{r, optimized for each area}
# ── Score each candidate seed per stratum separately ─────────────────────────

# Extract per-stratum centroid X position for each candidate realisation
stratum_positions <- lapply(region_split@strata.name, function(s) {
  sapply(valid_idx, function(i) {
    t <- candidate_transects[[i]]
    lines <- t@samplers[t@samplers$strata == s, ]
    
    if (nrow(lines) == 0) return(NA)
    
    centroids <- sf::st_centroid(sf::st_geometry(lines))
    coords    <- sf::st_coordinates(centroids)
    mean(coords[, 1])   # X position (easting) per stratum
  })
})
names(stratum_positions) <- region_split@strata.name

# Normalise each stratum's positions to 0-1 independently
stratum_norm <- lapply(stratum_positions, function(pos) {
  rng <- range(pos, na.rm = TRUE)
  if (diff(rng) == 0) return(rep(0.5, length(pos)))
  (pos - rng[1]) / diff(rng)
})

# ── Greedy selection using per-stratum spread scores ─────────────────────────
# At each step, pick the candidate that maximises the MINIMUM spread
# across all three strata — forces balanced coverage everywhere

selected_idx <- integer(n_survey_days)
remaining    <- seq_along(valid_idx)

# First pick: closest to median in ALL strata simultaneously
first_scores <- sapply(remaining, function(j) {
  mean(sapply(stratum_norm, function(sn) abs(sn[j] - 0.5)), na.rm = TRUE)
})
selected_idx[1] <- remaining[which.min(first_scores)]
remaining <- remaining[-which(remaining == selected_idx[1])]

for (k in 2:n_survey_days) {
  
  # For each remaining candidate, compute its minimum distance
  # to already-selected seeds, separately per stratum
  spread_scores <- sapply(remaining, function(j) {
    
    per_stratum_min_dist <- sapply(stratum_norm, function(sn) {
      already <- sn[selected_idx[1:(k-1)]]
      min(abs(sn[j] - already), na.rm = TRUE)
    })
    
    # Score = minimum spread across strata — weakest stratum drives selection
    min(per_stratum_min_dist, na.rm = TRUE)
  })
  
  best <- remaining[which.max(spread_scores)]
  selected_idx[k] <- best
  remaining <- remaining[-which(remaining == best)]
}

# ── Rebuild schedule with new seeds ──────────────────────────────────────────
selected_seeds <- candidate_seeds[valid_idx[selected_idx]]

survey_schedule <- data.frame(
  day       = 1:n_survey_days,
  seed      = selected_seeds,
  date      = NA,
  completed = FALSE,
  effort_km = NA,
  notes     = NA
)

cat("=== Updated Survey Schedule ===\n")
print(survey_schedule[, c("day", "seed")])

# ── Verify per-stratum spread visually ───────────────────────────────────────
# Show how evenly the selected seeds sample each stratum's positional range
cat("\n=== Per-stratum positional spread of selected seeds ===\n")
for (s in region_split@strata.name) {
  positions <- stratum_norm[[s]][selected_idx]
  cat("Stratum", s,
      "| Range covered:", round(diff(range(positions, na.rm=TRUE)), 3),
      "| SD:", round(sd(positions, na.rm=TRUE), 3), "\n")
}
# You want Range close to 1.0 and SD as even as possible across strata
# If one stratum has Range << 1.0, increase n_candidates and rerun

# ── Replot with new seeds ─────────────────────────────────────────────────────
transect_df <- lapply(1:n_survey_days, function(k) {
  i  <- valid_idx[selected_idx[k]]
  t  <- candidate_transects[[i]]
  df <- sf::st_sf(geometry = sf::st_geometry(t@samplers))
  df$day     <- paste0("Day ", sprintf("%02d", k))
  df$strata  <- t@samplers$strata
  df
}) |> dplyr::bind_rows()

# Facet plot coloured by stratum so you can judge each area separately
ggplot() +
  geom_sf(data      = sf::st_sf(geometry = sf::st_geometry(region_split@region)),
          colour = "#1e3a5f", linewidth = 0.4) +
  geom_sf(data      = transect_df,
          aes(colour = strata),
          linewidth = 0.5, alpha = 0.85) +
  scale_colour_manual(
    values = c("A" = "#38bdf8", "B" = "#f59e0b", "C" = "#10b981"),
    name   = "Stratum"
  ) +
  facet_wrap(~day, ncol = 5) +
  labs(title    = "Survey Transects — Per-stratum Optimised Seeds",
       subtitle = "Each stratum's shift is independently maximised across days",
       caption  = paste0("Transect Lenght: ", paste(round(sum(candidate_transects[[1]]@line.length)/1000, 1), "km"))
       ) +
  theme_minimal(base_size = 10)


transect_df %>% 
  filter(day == "Day 01") %>% 
  select(geometry) %>% 
  plot()

```



# Adjusting spacing again for feasible on effort hours

```{r}

# ── Parameters ────────────────────────────────────────────────────────────────
vessel_speed_kmh <- 15      # on-effort speed (~6.5 knots)
max_hours        <- 5.0     # maximum feasible hours per day
target_coverage  <- 0.85    # slightly relaxed — discuss with supervisor
truncation_m     <- 300     # design truncation (field), analysis can differ

max_effort_km    <- vessel_speed_kmh * max_hours
cat("Max effort per day:", max_effort_km, "km\n\n")

# ── Test spacings ─────────────────────────────────────────────────────────────
spacings <- seq(700, 1500, by = 100)

results <- data.frame(
  spacing      = spacings,
  expected_cov = round((2 * truncation_m) / spacings, 3),
  effort_km    = NA,
  hours        = NA,
  n_transects  = NA,
  mean_cov     = NA,
  cv           = NA,
  feasible     = NA
)

for (i in seq_along(spacings)) {
  d <- make.design(
    region        = region_split,
    transect.type = "line",
    design        = "eszigzag",
    spacing       = rep(spacings[i], 3),
    design.angle  = c(90, 0, 0),
    edge.protocol = "minus",
    truncation    = truncation_m
  )
  
  t      <- generate.transects(d)
  effort <- sum(as.numeric(sf::st_length(t@samplers))) / 1000
  hours  <- effort / vessel_speed_kmh
  
  if (effort > 0) {
    cov    <- run.coverage(d, reps = 15)
    scores <- unlist(cov@coverage.scores)
    
    results$effort_km[i]   <- round(effort, 1)
    results$hours[i]       <- round(hours, 2)
    results$n_transects[i] <- nrow(t@samplers)
    results$mean_cov[i]    <- round(mean(scores), 3)
    results$cv[i]          <- round(sd(scores)/mean(scores), 3)
    results$feasible[i]    <- ifelse(
      hours <= max_hours & mean(scores) >= target_coverage, "✓", "✗"
    )
  }
  
  cat("Spacing", spacings[i], "| Effort:", round(effort,1), "km |",
      round(hours, 1), "h | Coverage:", 
      ifelse(effort > 0, round(mean(unlist(cov@coverage.scores)), 3), "fail"), "\n")
}

print(results)



```


Results:

 spacing expected_cov effort_km hours n_transects mean_cov    cv feasible
1     700        0.857     118.3  7.89          29    0.838 0.265        ✗
2     800        0.750     105.4  7.03          26    0.735 0.289        ✗
3     900        0.667      92.4  6.16          23    0.649 0.227        ✗
4    1000        0.600      85.3  5.68          22    0.598 0.269        ✗
5    1100        0.545      77.5  5.17          20    0.541 0.253        ✗
6    1200        0.500      71.4  4.76          18    0.496 0.286        ✗
7    1300        0.462      69.1  4.61          18    0.462 0.286        ✗
8    1400        0.429      62.1  4.14          17    0.430 0.291        ✗
9    1500        0.400      58.7  3.91          16    0.405 0.400        ✗



## Run cummulative days with shifted starting points 


```{r}

simulate_cumulative_coverage <- function(design, n_days, n_sims = 200,
                                         grid_size = 5000) {
  
  # Get region geometry for grid generation
  region_bbox <- sf::st_bbox(region_split@region)
  
  # Generate a coverage grid
  grid_pts <- sf::st_sample(
    sf::st_as_sfc(region_bbox),
    size = grid_size,
    type = "regular"
  )
  grid_pts <- grid_pts[sf::st_within(grid_pts,
                        sf::st_union(region_split@region),
                        sparse = FALSE)[,1]]
  
  cat("Coverage grid:", length(grid_pts), "points\n")
  
  # For each simulation, stack n_days realisations
  coverage_by_day <- matrix(NA, nrow = n_sims, ncol = n_days)
  
  for (sim in 1:n_sims) {
    if (sim %% 50 == 0) cat("Simulation", sim, "/", n_sims, "\n")
    
    # Track which grid points have been covered
    covered <- rep(FALSE, length(grid_pts))
    
    for (day in 1:n_days) {
      # New random transect realisation each day
      t <- generate.transects(design)
      lines <- sf::st_geometry(t@samplers)
      
      # Buffer each line by truncation distance
      lines_buffered <- sf::st_buffer(lines, dist = truncation_m)
      union_buffer   <- sf::st_union(lines_buffered)
      
      # Update which points are now covered
      newly_covered <- sf::st_within(grid_pts, union_buffer,
                                      sparse = FALSE)[,1]
      covered <- covered | newly_covered
      
      # Coverage = proportion of region covered at least once
      coverage_by_day[sim, day] <- mean(covered)
    }
  }
  
  # Summarise across simulations
  data.frame(
    day      = 1:n_days,
    mean_cov = round(colMeans(coverage_by_day), 3),
    lower_95 = round(apply(coverage_by_day, 2,
                            quantile, probs = 0.025), 3),
    upper_95 = round(apply(coverage_by_day, 2,
                            quantile, probs = 0.975), 3),
    sd       = round(apply(coverage_by_day, 2, sd), 4)
  )
}

# ── Run for your wide-spacing design ─────────────────────────────────────────
cat("Simulating cumulative coverage...\n")
cum_cov <- simulate_cumulative_coverage(
  design  = d,
  n_days  = 15,
  n_sims  = 100
)

print(cum_cov)


```

Results:

Used design with 1500 m of spacing:

   day mean_cov lower_95 upper_95     sd
1    1    0.390    0.373    0.406 0.0089
2    2    0.628    0.525    0.688 0.0497
3    3    0.758    0.599    0.841 0.0568
4    4    0.847    0.741    0.925 0.0503
5    5    0.905    0.818    0.964 0.0383  -> Point with almost 100 % coverage
6    6    0.939    0.863    0.986 0.0319
7    7    0.963    0.911    0.994 0.0239
8    8    0.977    0.936    0.997 0.0189
9    9    0.984    0.948    0.998 0.0167  
10  10    0.991    0.963    1.000 0.0097
11  11    0.994    0.976    1.000 0.0068
12  12    0.996    0.979    1.000 0.0060
13  13    0.997    0.983    1.000 0.0051
14  14    0.998    0.988    1.000 0.0036
15  15    0.999    0.994    1.000 0.0021



```{r}

cum_cov %>% 
  ggplot(aes(
    x = day,
    y = mean_cov
  )) +
  geom_point(colour = "darkblue", size = 5) +
  geom_line(colour = "darkblue", linewidth = 3)+
  geom_hline(
    yintercept = .9,
    linetype =  "dashed",
    linewidth = 2,
    alpha = 0.3
  ) +
  geom_vline(
    xintercept = 5,
    linetype =  "dotted",
    linewidth = 2,
    alpha = 0.3
  ) +
  annotate("text", x = 2, y = .92, label = "90 % coverage line", alpha = 0.3)+
  
  scale_x_continuous(breaks = cum_cov$day) +
  
  
  theme_minimal(base_size =  15)



```



# Compare covered area

Calculate the covered area from line length * 2* truncation and use geometry to account for turning and areas outside the study area

```{r}

truncation_m = 300



strips <- sf::st_buffer(
  sf::st_geometry(transects_c@samplers),
  dist = truncation_m
)

# Union all strips into one polygon (merges overlapping areas at turns)
covered_polygon <- sf::st_union(strips)

plot(covered_polygon)

# Clip to study region so you don't count area outside your boundary
covered_clipped <- sf::st_intersection(
  covered_polygon,
  sf::st_union(region_split@region)
)

plot(covered_clipped)

region_split@region

# True covered area (no double-counting of overlapping strips)
true_covered_km2 <- as.numeric(sf::st_area(covered_clipped)) / 1e6
cat("True covered area (clipped, no overlap):", 
    round(true_covered_km2, 2), "km²\n")


ggplot() +
  geom_sf(data      = sf::st_sf(geometry = sf::st_geometry(region_split@region)),
          colour = "#1e3a5f", linewidth = 0.4) +
  geom_sf(data = sf::st_sf(geometry = sf::st_geometry(covered_clipped)),
          fill = "darkred", alpha = .5)




(true_covered_km2/(sum(region_split@area)/1e6))

```

