---
title: "Visualisation"
format: html
editor: source
---

```{r}
library(tidyverse)
library(readxl)
library(lubridate)
library(sf)
library(ggspatial)
library(osmdata)
library(ggforce)
library(gt)
library(ggtext)
library(legendry)
library(ks)  # Kernel smoothing for Kernel density Map
library(patchwork)
library(flextable)
library(ggfun)
library(ggsignif)
library(ggmagnify)
library(tinytable)
library(ggmap)


```

#### Colorpalette

```{r}
darkblue = "#212842"
midblue = "#4E5B9C"
lightblue = "#5C9EAD"

```

```{r}
# Excel sheet of the spots with details
spots <- read_excel(path = "./data/datasheet_landsurvey.xlsx", sheet = "spots")


# Gib Bay coastline
coast <- st_read("./data/bayofalgib.gpkg")

# Clean the sighting data and import
source("./land_survey.R")


# Set active spots as variable
active_spots <- spots %>%  
  filter(spot_name != "MedSteps" & spot_name != "WestSide") %>% 
  select(spot_name)

```

# Create a map of the survey sectors

This maps shows the sectors that were surveyed.

```{r}
make_wedge <- function(x0, y0, r, start, end, crs, n = 120) {
  b   <- seq(start, end, length.out = n) * pi / 180
  pts <- cbind(x0 + r * sin(b), y0 + r * cos(b))   # compass: x=sin, y=cos
  ring <- rbind(c(x0, y0), pts, c(x0, y0))         # close through the centre
  st_sfc(st_polygon(list(ring)), crs = crs)
}

proj <- 32630   # UTM zone 30N — metres, correct for Gibraltar

R = 6.378e+6


centres <- as.data.frame(spots)
centres <- centres %>% 
  rename(name = spot_name,
         start = spot_bearing1,
         end = spot_bearing2) %>% 
  mutate(r_m = sqrt(2*R*observer_height/20))


# project the centres into metres
xy <- st_coordinates(
  st_transform(st_as_sf(centres, coords = c("lon", "lat"), crs = 4326), proj)
)

sectors <- do.call(rbind, lapply(seq_len(nrow(centres)), function(i) {
  g <- make_wedge(xy[i, 1], xy[i, 2], centres$r_m[i],
                  centres$start[i], centres$end[i], crs = proj)
  st_sf(name = centres$name[i], geometry = g)
}))

sectors <- st_transform(sectors, 4326)   # back to lon/lat for the map

sectors <- sectors  %>% 
  filter(name != "MedSteps" & name != "WestSide")

labels <- centres %>%
  filter(name != "MedSteps" & name != "WestSide") %>% 
  mutate(
    nudge_x = c(0.00, 0.00, 0.00),
    nudge_y = c(-0.006, 0.005, 0.005),
    fill = c(darkblue, lightblue, midblue)
  )
labels <- st_transform(st_as_sf(labels, coords = c("lon", "lat"), crs = 4326), proj)

# Sector map
sector_map <- ggplot() +
  geom_sf(data = coast) +
  geom_sf(data = sectors, aes(fill = name), colour = "black", alpha = 0.6) +
  geom_sf(data = st_set_geometry(data, "geometry") %>% filter(spot_name %in% active_spots$spot_name), shape = 25, fill = "gray10") +
  geom_sf_label(
    aes(label = name),
    data = labels,
    nudge_x = labels$nudge_x,
    nudge_y = labels$nudge_y,
    fill = labels$fill,
    color = "white")+
  coord_sf(
    xlim = c(-5.438, -5.248),
    ylim = c(36.05, 36.195)
  ) +
  
  labs(fill = "") +
  
  # Add the scale bar
  annotation_scale(
    location = "bl",            # "bl" = bottom left
    width_hint = 0.2,          # Bar length takes up approx 40% of map width
    unit_category = "metric",   # Options: "metric", "imperial", or "nautical"
    pad_y = unit(0.05, "npc")
  ) +
  
  # Add the north arrow
  annotation_north_arrow(
    location = "br", # "tr" = top right
    height = unit(0.12, "npc"),
    width = unit(0.1, "npc"),
    which_north = "true",       # Align arrow to true north pole instead of map grid
    pad_x = unit(0.05, "npc"),    # X offset from the boundary
    pad_y = unit(0.05, "npc"),    # Y offset from the boundary
    style = north_arrow_nautical # Arrow graphic style
  ) +
  
  scale_fill_manual(values = c(
    Europa = darkblue,
    PrinceGeorge = midblue,
    SandyBay = lightblue
  )) +
  my_theme_trans() +
  theme(legend.position = "none")

sector_map


```

Overview map, with inset

First create inset
```{r}

europe <- map_data("world") %>% 
  filter(
    long >= -25,
    long <= 45,
    lat >= 30,
    lat <= 72
  )

inset <- ggplot(europe, aes(long, lat, group = group)) +
  geom_polygon(fill = "lightgrey", color = "black", linewidth = 0.2) +
  geom_rect(
    aes(
      xmin = -5.438,
      xmax = -5.248,
      ymin = 36.05,
      ymax = 36.195
    ),
    color = "red",
    fill = NA,
    linewidth = 1
  ) +
  annotate(
    "label",
    x = -8,
    y = 35.9,
    label = "Strait of Gibraltar",
    size = 2.5,
    fontface = "italic"
  ) +
  annotate(
    "text",
    x = -3,
    y = 38.950,
    label = "Spain",
    size = 5
  ) +
  coord_fixed(
    xlim = c(-10, 0),
    ylim = c(35, 40)
  ) +
  theme_void(base_size = 5) +
  theme(panel.border = element_rect(fill = NA, colour = "black", linewidth = 0.5),
        panel.background = element_rect(fill = "white"))
```

Insert the inset with `patchwork`

```{r}

overview <- sector_map + inset_element(inset,left = 0.00001, bottom = 0.8, right = 0.3, top = 1)

overview
```



```{r, Export}
ggsave(filename = "./figures/sector_map.png", sector_map, width = 1920, height = 1920, dpi = 200, units = "px")

ggsave(filename = "./figures/overview_map.png", overview, width = 1920, height = 1920, dpi = 200, units = "px")

```

### Create overview map with satelite map 

```{r}

```


# Visualise the sightings

::: callout-warning
Old script
:::

```{r}

# Create labels for the different spots
spot_label <- data %>% 
  select(spot_name, geometry, observer_height) %>% 
  mutate(y_nudge = case_when(
    spot_name == "SandyBay" ~ 0.003,
    spot_name == "Europa" ~ 0.003,
    spot_name == "WestSide" ~ 0.003,
    
    TRUE ~ - 0.003),
    
    x_nudge = case_when(
      spot_name == "WestSide" ~ -0.002,
      spot_name == "MedSteps" ~ 0.002,
      
      TRUE ~ 0
    )
  )


sightings <- ggplot() +
  geom_sf(data = coast, fill = "lightgray", alpha = .8) +
  geom_sf(data = st_set_geometry(data, "geometry")) +   # Set geometry so aes does not missalign
  # geom_sf(data = sectors, aes(fill = name), colour = "black", alpha = 0.4) +
  geom_sf(data = st_set_geometry(data, "sighting_poi"), #%>%  filter(date.x >= "2026-06-09"),
          aes(fill = species, linetype = species ), colour = "black", size = 3, shape = 21) +
  # geom_sf_label(data = st_set_geometry(data, "geometry"),
  #               aes(label = spot_name), size = 2, nudge_y = spot_label$y_nudge) +
  # geom_sf_text(data = st_set_geometry(data, "sighting_poi"),
  #             aes(label = species), size = 3, nudge_y = -0.003) +
  #facet_wrap(~ sector) +
  
  
  coord_sf(
    xlim = c(-5.438, -5.248),
    ylim = c(36.05, 36.195)
  ) +
  labs(fill = "Species",
       linetype = "Whales") +
  theme_minimal() +
  
  scale_shape_manual(values = c(21, 22, 24)) + 
  
  scale_fill_manual(values = c(     # If only one value is defined the rest is grayed out by default
   delphinidae = "lightblue",
   common_dolphin = "darkblue",
   striped_dolphin = "blue",
   fin_whale = "purple"
   
  ),
  labels = c(
     delphinidae = "Delphinidae",
     common_dolphin = "<b>Common Dolphin</b> <br> <i>Delphinus delphis</i>",
     striped_dolphin = "Striped Dolphin",
     fin_whale = "<b>Fin whale</b> <br> <i> Balaenoptera physalus </i>"
     
   ))+
  
  my_theme_trans() +
  
  theme(
  legend.text = ggtext::element_markdown()
)

sightings


```

### Include the boat data and combine to one df

```{r}

# Get the boat data from the data cleaning script
boat <- readRDS("./temp/boat_data.rds")

boat <- boat %>% 
  select(date.x, species, group_size, behaviour, geometry) %>% 
  mutate(
    source = "boat"
  )

land <- data %>% 
  select(date.x, species, group_size, behaviour, sighting_poi) %>% 
  rename(geometry = sighting_poi) %>% 
  mutate(
    source = "land"
  )

full <- boat %>%
  bind_rows(land)

saveRDS(full, "./temp/full.rds")

full_source <- full %>% 
  summarise(
    sightings = n(),
    .by = "source"
  ) %>% 
  pivot_wider(
    names_from = source,
    values_from = sightings
  )

st_set_geometry(full, "geometry")

full_sum <- full %>% 
  summarise(
    sightings = n()
  )

#--------------

mt_levels = c("boat", "land")
mt_groups = c("Method", "Method")


mt_fill = c(darkblue, lightblue)

mt_fill_labs = c(paste0("Boat Transects (n = ", full_source$boat, ")"), paste0("Land Surveys (n = ", full_source$land, ")"))

mt_headers = paste0(mt_groups, " (n = ", full_sum$sightings, ")")



# Plot
method_map <- ggplot() +
  geom_sf(data = coast, fill = "lightgray", alpha = .8) +
  
  geom_sf(data = st_set_geometry(full, "geometry"),
          aes(fill = source, shape = source), colour = "black", size = 3) + 
  coord_sf(xlim = c(-5.47, -5.248), ylim = c(36.05, 36.2005)) +
  
  scale_shape_manual(
    values = c(22, 21),
    labels = mt_fill_labs,
    name = ""
  ) +
  
  scale_fill_manual(
    values = mt_fill,
    labels = mt_fill_labs,
    name = ""
  ) +
  
  
  
  # Add the scale bar
  annotation_scale(
    location = "bl",            # "bl" = bottom left
    width_hint = 0.2,          # Bar length takes up approx 40% of map width
    unit_category = "metric",   # Options: "metric", "imperial", or "nautical"
    pad_y = unit(0.05, "npc")
  ) +
  
  # Add the north arrow
  annotation_north_arrow(
    location = "br", # "tr" = top right
    height = unit(0.12, "npc"),
    width = unit(0.1, "npc"),
    which_north = "true",       # Align arrow to true north pole instead of map grid
    pad_x = unit(0.05, "npc"),    # X offset from the boundary
    pad_y = unit(0.05, "npc"),    # Y offset from the boundary
    style = north_arrow_nautical # Arrow graphic style
  ) +
  
  guides(fill = guide_legend_group(key = key_group_lut(mt_levels, mt_headers),
                                   override.aes = list(colour = "black")),
         shape = guide_legend_group(key = key_group_lut(mt_levels, mt_headers),
                                   override.aes = list(colour = "black"))) +
  
  theme_minimal() +
  my_theme_trans() +
  theme(
    legend.text              = ggtext::element_markdown(),
    legendry.legend.subtitle = element_text(face = "bold"),   # styles the group headers
    legend.position = "inside",
    legend.position.inside = c(0.12, 0.92)
  )

method_map


```

```{r}
ggsave(filename = "./figures/method_map.png", method_map, width = 3020, height = 2520, dpi = 300, units = "px")
```




Nesting the legend to make it easier understandable

```{r}

# species & their groups, in display order
sp_levels <- c("common_dolphin", "striped_dolphin", "delphinidae", "fin_whale", "unidentified")
sp_groups <- c("Odontoceti",     "Odontoceti",      "Odontoceti",  "Mysticeti", "Other")

# count sightings per species, then sum per group
sp_counts  <- table(factor(data$species, levels = sp_levels))   # per-species counts
grp_totals <- tapply(as.integer(sp_counts), sp_groups, sum)     # summed per group

# bake the totals into the group-header strings
group_headers <- paste0(sp_groups, " (n = ", grp_totals[sp_groups], ")")

species_groups <- key_group_lut(sp_levels, group_headers)


# Define shape color and legend entry once
data$species <- factor(data$species, levels = sp_levels)

# define the key contents ONCE so fill and shape stay identical
sp_labels <- c(
  common_dolphin  = "<b>Common Dolphin</b><br><i>Delphinus delphis</i>",
  striped_dolphin = "<b>Striped Dolphin</b><br><i>Stenella coeruleoalba</i>",
  delphinidae     = "<b>Delphinidae</b>",
  fin_whale       = "<b>Fin whale</b><br><i>Balaenoptera physalus</i>",
  unidentified    = "Unidentified"
)
sp_shapes <- c(common_dolphin = 21, striped_dolphin = 21, delphinidae = 21,
               fin_whale = 22, unidentified = 21)
sp_fills  <- c(common_dolphin = darkblue, striped_dolphin = midblue, delphinidae = lightblue,
               fin_whale = "purple", unidentified = "gray")


# Plot
sightings <- ggplot() +
  geom_sf(data = coast, fill = "lightgray", alpha = .8) +
  geom_sf(data = st_set_geometry(data, "geometry") %>% filter(spot_name %in% active_spots$spot_name), shape = 25, fill = "gray10", size = 3.2) +
  geom_sf(data = st_set_geometry(data, "sighting_poi"),
          aes(fill = species, shape = species), colour = "black", size = 3) + 
  coord_sf(xlim = c(-5.50, -5.248), ylim = c(36.05, 36.2005)) +
  
  scale_shape_manual(
    values = sp_shapes,
    labels = sp_labels,
    name = ""
  ) +
  
  scale_fill_manual(
    values = sp_fills,
    labels = sp_labels,
    name = ""
  ) +
  
  # Add the scale bar
  annotation_scale(
    location = "bl",            # "bl" = bottom left
    width_hint = 0.2,          # Bar length takes up approx 40% of map width
    unit_category = "metric",   # Options: "metric", "imperial", or "nautical"
    pad_y = unit(0.05, "npc")
  ) +
  
  # Add the north arrow
  annotation_north_arrow(
    location = "br", # "tr" = top right
    height = unit(0.12, "npc"),
    width = unit(0.1, "npc"),
    which_north = "true",       # Align arrow to true north pole instead of map grid
    pad_x = unit(0.05, "npc"),    # X offset from the boundary
    pad_y = unit(0.05, "npc"),    # Y offset from the boundary
    style = north_arrow_nautical # Arrow graphic style
  ) +
  
  guides(fill = guide_legend_group(key = species_groups,
                                   override.aes = list(colour = "black")),
         shape = guide_legend_group(key = species_groups,
                                   override.aes = list(colour = "black"))) +   # <- the nesting happens here
  theme_minimal() +
  my_theme_trans() +
  theme(
    legend.text              = ggtext::element_markdown(),
    legendry.legend.subtitle = element_text(face = "bold"),   # styles the group headers
    legend.position = "inside",
    legend.position.inside = c(0.15, 0.7)
  )

sightings


```

```{r, Export}
ggsave(filename = "./figures/sightings_legend.png", sightings, width = 3020, height = 2320, dpi = 300, units = "px")
```

#### Pull Obis data and compare

```{r}
library(robis)

xlim <- c(-5.438, -5.248)
ylim <- c(36.05, 36.195)


wkt <- sprintf(
  "POLYGON ((%f %f, %f %f, %f %f, %f %f, %f %f))",
  xlim[1], ylim[1],   # SW
  xlim[2], ylim[1],   # SE
  xlim[2], ylim[2],   # NE
  xlim[1], ylim[2],   # NW
  xlim[1], ylim[1]    # close back to SW
)

obis_dd <- occurrence("Delphinus delphis", geometry = wkt)

obis_sf <- st_as_sf(obis_dd,
                    coords = c("decimalLongitude", "decimalLatitude"),
                    crs = 4326,
                    na.fail = FALSE)


```

Compare to my sightings

```{r}
sightings <- data %>%
  filter(on_effort == TRUE & date.x > "2026-05-28") %>% 
  st_set_geometry("sighting_poi") %>%
  st_transform(4326) %>%
  transmute(source = "This study",
            geometry = sighting_poi)

obis <- obis_sf %>%
  st_transform(4326) %>%
  transmute(source = "OBIS")

combined <- bind_rows(sightings, obis)
combined$source <- factor(combined$source,
                          levels = c("This study", "OBIS"))

key_labs = c(paste0("This Study (n = ", length(sightings$source), ")"), paste0("OBIS (n = ", length(obis$source), ")"))


obis_comp <- obis_sf %>% 
  ggplot() +
  geom_sf(data = coast, fill = "lightgray", alpha = .8) +
  geom_sf(data = st_set_geometry(combined, "geometry"), aes(
    shape = source, fill = source, alpha = source), colour = "black", size = 3) +
  scale_shape_manual(values = c("This study" = 21,
                                "OBIS" = 22),
                     labels = key_labs) +
  
  
  scale_fill_manual(values  = c("This study" = lightblue,
                                "OBIS" = darkblue),
                    labels = key_labs) +
  
  
  scale_alpha_manual(values  = c("This study" = 0.3,
                                "OBIS" = 1),
                     labels = key_labs) +
  
  
  coord_sf(xlim = c(-5.50, -5.248), ylim = c(36.05, 36.2005)) +
  labs(shape = "Source", fill = "Source", alpha = "Source") +
  
  # Add the scale bar
  annotation_scale(
    location = "bl",            # "bl" = bottom left
    width_hint = 0.2,          # Bar length takes up approx 40% of map width
    unit_category = "metric",   # Options: "metric", "imperial", or "nautical"
    pad_y = unit(0.05, "npc")
  ) +
  
  # Add the north arrow
  annotation_north_arrow(
    location = "br", # "tr" = top right
    height = unit(0.12, "npc"),
    width = unit(0.1, "npc"),
    which_north = "true",       # Align arrow to true north pole instead of map grid
    pad_x = unit(0.05, "npc"),    # X offset from the boundary
    pad_y = unit(0.05, "npc"),    # Y offset from the boundary
    style = north_arrow_nautical # Arrow graphic style
  ) +
  
  my_theme_trans() +
  theme(
    legend.text              = ggtext::element_markdown(),
    legendry.legend.subtitle = element_text(face = "bold"),   # styles the group headers
    legend.position = "inside",
    legend.position.inside = c(0.10, 0.7)
  )
```

```{r}
ggsave(filename = "./figures/obis_comp.png", obis_comp, width = 3020, height = 2320, dpi = 300, units = "px") 
```

```{r}
station_to_sector <- c(
  Europa       = "S",
  PrinceGeorge = "W",
  SandyBay     = "E"
)

obis_sf <- st_transform(obis_sf, st_crs(sectors))

obis_counts <- st_join(obis_sf, sectors, join = st_within) %>%
  st_drop_geometry() %>%
  count(name, name = "records") %>%     # use sectors' column name here
  mutate(dataset = "OBIS")

sys_counts <- data %>%
  st_drop_geometry() %>%
  filter(on_effort) %>%
  count(sector, name = "records") %>%
  mutate(dataset = "This study")

bind_rows(sys_counts, obis_counts) %>%
  tidyr::pivot_wider(names_from = sector, values_from = records, values_fill = 0)

obis_counts2 <- obis_counts %>%
  mutate(sector = recode(name, !!!station_to_sector)) %>%
  select(sector, records, dataset)   # sector is NA for the 45 outside-polygon records

comparison <- bind_rows(sys_counts, obis_counts2) %>%
  mutate(sector = factor(sector, levels = c("E", "S", "W"),
                         labels = c("East", "South", "West"))) %>%
  tidyr::pivot_wider(names_from = sector, values_from = records,
                     values_fill = 0)

comparison

saveRDS(comparison, "./temp/obis.rds")

```


# Sightings analysis

```{r}
# Filter out sightings before 28.05.2026
data_an <- data %>% filter(date.x > "2026-05-28")

data_an %>% 
  filter(on_effort == TRUE) %>% 
  summarise(n_sightings = n())

data_an %>% 
  filter(species %in% c("delphinidae", "common_dolphin")) %>% 
  summarise(.by = sector, n = n())

data_an %>% 
  filter(species %in% c("delphinidae", "common_dolphin")) %>% 
  summarise(.by = species, n = n())

# ----------------
# Analyse effort

effort <- effort_log %>% 
  mutate(date = dmy(date)) %>% 
  #filter(date > "2026-05-28") %>% 
  mutate(across(c(observation_start, observation_end),
                ~ hms::as_hms(.))) %>% 
  filter(!grepl("off_effort", survey_id)) %>% 
  mutate(effort_time = as.duration(observation_end - observation_start))

cat("Watch sessions:", nrow(effort),
    "| total on-effort hours:", round(sum(effort$effort_time)/60^2, 2), "\n")





cat("\nEncounter rate by station:\n"); print(er)


effort %>%  
  as.data.frame() %>% 
  mutate(
    dur = as.duration(observation_end - observation_start)
  ) %>% 
  select(survey_id, spot_name, dur) %>% 
  summarise(
    n = n(),
    .by = spot_name
  )

effort %>% 
  as.data.frame() %>%
  mutate(
    dur = as.duration(observation_end - observation_start)
  ) %>% 
  select(survey_id, spot_name, dur) %>% 
  summarise(
    time = sum(dur),
    .by = spot_name
  )


surveys_spot <- data_an %>% 
  summarise(
    n = n(),
    .by = spot_name
  )



# On effort # On effort # On effort hours per person
per_person <- effort_log %>% 
  mutate(duration = as.numeric(
    difftime(ymd_hms(observation_end), ymd_hms(observation_start), units = "mins")
  )) %>% 
  separate_longer_delim(crew, delim = ";") %>% 
  mutate(crew = trimws(crew)) %>% 
  summarise(
    total_mins = sum(duration, na.rm = TRUE),
    total_hours = sum(duration, na.rm = TRUE)/60,
    .by = "crew"
  )

per_person

# Watches per day
as.data.frame(effort_log) %>% 
  mutate(date = dmy(date)) %>% 
  summarise(
    n = n(),
    .by = "date"
  ) %>% 
  
  ggplot(aes(x = date, y = n)) +
  geom_col(fill = darkblue) +
  scale_x_date(
    date_breaks = "2 days",
    date_labels = "%d %b"
  ) +
  my_theme_trans() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r, sightings time}
# When did we have the most sightings?
sight_time <- data %>%
  mutate(
    time_bin = paste0(
      sprintf("%02d", hour(start_time)),
      ":",
      ifelse(minute(start_time) < 30, "00", "30")
    )
  ) %>%
  count(time_bin)
 
sighting_day <- sight_time %>% 
  ggplot(aes(x = time_bin, y = n)) +
  geom_col(fill = darkblue) +
  my_theme_trans() +
  labs(
    x = "Time",
    y = "n of Sightings",
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

sighting_day
  
```

```{r}
ggsave(filename = "./figures/sighting_day.png", sighting_day, width = 1920, height = 1520, dpi = 200, units = "px")
```

### Does tide have an influence? 

```{r}
tide_df <- data %>% 
  mutate(tide = as.factor(tide)) %>% 
  summarise(
    n = n(),
    .by = "tide"
  ) %>% 
  ggplot(aes(x = tide, y = n)) +
  geom_col()

tide_df

```


### Compare current to target watches/hours
Target hours comes from the Power Analysis script


```{r}
per_spot <- as.data.frame(effort) %>% 
  filter(spot_name %in% active_spots$spot_name) %>% 
  # mutate(duration = as.numeric(
  #   difftime(ymd_hms(observation_end), ymd_hms(observation_start), units = "mins")
  # )) %>% 
  summarise(
    total_hours = sum(effort_time, na.rm = TRUE)/60^2,
    total_mins = total_hours*60,
    percent = round((total_hours / 9) * 100, 2), 
    .by = "spot_name"
  )

aim <- data.frame(
  spot = per_spot$spot_name,
  target = 9
)

hour_spot <- per_spot %>% 
  ggplot(aes(x = spot_name, y = per_spot$total_hours)) +
  geom_col(fill = darkblue) +
  geom_text(aes(x = spot_name, y = total_hours + 0.5), ,label = paste(per_spot$percent, "%")) +
  geom_errorbar(
    data = aim,
    aes(x = spot, ymin = target, ymax = target),
    linewidth = 1.5,
    linetype = "dashed",
    color = "darkred"
  ) +
  labs(
    x = "",
    y = "Hours per spot"
  ) +
  my_theme_trans()
  
hour_spot
```

```{r}
ggsave(filename = "./figures/hour_spot.png", hour_spot, width = 1920, height = 1520, dpi = 200, units = "px")
```


### Convert to sightings per unit effort (SPUE)

```{r}
effort <- effort_log %>%
  filter(!str_detect(survey_id, "off_effort")) %>%
  mutate(
    duration_min = as.numeric(difftime(observation_end, observation_start,
                                        units = "mins"))
  )
 
# ---- 3. Detections: keep on-effort sightings -------------------------------
sightings <- detection %>%
  filter(on_effort == 1)

 
# ---- 4. Sightings & individuals per scan -----------------------------------
per_scan <- sightings %>%
  group_by(survey_id) %>%
  summarise(
    n_sightings   = n(),
    n_individuals = sum(group_size, na.rm = TRUE),  # one NA group size -> 0
    .groups = "drop"
  )
 
# Join back onto the full effort table; scans with no sighting become 0 (NOT NA)
effort <- effort %>%
  left_join(per_scan, by = "survey_id") %>%
  mutate(
    n_sightings   = replace_na(n_sightings, 0),
    n_individuals = replace_na(n_individuals, 0)
  )
 
# ---- 5. SPUE by spot -------------------------------------------------------
spue_by_spot <- effort %>%
  group_by(spot_name) %>%
  summarise(
    scans        = n(),
    effort_hours = sum(duration_min, na.rm = TRUE) / 60,
    sightings    = sum(n_sightings),
    individuals  = sum(n_individuals),
    .groups = "drop"
  ) %>%
  mutate(
    spue_per_scan   = sightings / scans,        # headline SPUE
    sightings_per_h = sightings / effort_hours, # encounter rate
    indiv_per_h     = individuals / effort_hours
  ) %>%
  arrange(desc(scans))
 
# ---- 6. SPUE overall -------------------------------------------------------
spue_overall <- effort %>%
  summarise(
    spot_name    = "TOTAL",
    scans        = n(),
    effort_hours = sum(duration_min, na.rm = TRUE) / 60,
    sightings    = sum(n_sightings),
    individuals  = sum(n_individuals)
  ) %>%
  mutate(
    spue_per_scan   = sightings / scans,
    sightings_per_h = sightings / effort_hours,
    indiv_per_h     = individuals / effort_hours
  )
 
spue_summary <- bind_rows(spue_by_spot, spue_overall)
 
# ---- 7. Report & export ----------------------------------------------------
spue_report <- spue_summary %>%
  mutate(across(c(effort_hours, spue_per_scan, sightings_per_h, indiv_per_h),
                ~ round(.x, 3)))
 
print(spue_report, n = Inf)

spue_report %>% 
  filter(spot_name %in% active_spots$spot_name | spot_name == "TOTAL") %>% 
  select(!geometry) %>% 
  flextable() %>% 
  delete_columns(j = "geometry") %>% 
  set_header_labels(
    spot_name = "Spot Names",
    scans = "Scans",
    effort_hours = "Hours",
    sightings = "Sightings",
    individuals = "Individuals",
    spue_per_scan = "SPUE (per scan)",
    sightings_per_h = "Sightings per h",
    indiv_per_h = "Indiv. per h"
  ) %>% 
  colformat_double(
    j = c("effort_hours", "sightings_per_h", "sightings_per_h"),
    digits = 2
  ) %>% 
  bold(i = 4) %>% 
  autofit() %>% 
  theme_booktabs()

```

Analize spue and compare. Test if there is significant difference between the spots (Kruskal-Wallis) and between which spots (dun-Test). 

```{r}
library(FSA)

active <- c("Europa", "SandyBay", "PrinceGeorge")
scans  <- effort %>% filter(spot_name %in% active)   # per-scan rows, NOT the summary

#count(scans, spot_name)     


kruskal.test(n_sightings ~ spot_name, data = scans)

dunnTest(n_sightings ~ factor(spot_name), data = scans, method = "bh")
```

```{r}


spue_clean <-spue_report %>% 
  filter(spot_name %in% active_spots$spot_name) %>% 
  select(!geometry)

SPUE <- spue_clean %>% 
  ggplot(aes(x = spot_name, y = spue_per_scan))+
  geom_col(fill = darkblue) +
  geom_signif(
    # Specify which columns to connect (using group names or numeric X-indices)
    comparisons = list(c("Europa", "PrinceGeorge")),
    
    # Enter your exact pre-calculated p-value symbols in the same order
    annotations = c("***"), 
    map_signif_level = TRUE,
   
    
    # Visual styling details
    tip_length = 0.02,
    textsize = 6,
    
    
  ) +
  
  geom_signif(
    comparisons = list(c("Europa", "SandyBay")),
    annotations = c("*"),
    map_signif_level = TRUE,


    # Visual styling details
    tip_length = 0.02,
    textsize = 6,
    y_position=1.18
  ) +
  
  labs(x = "", y = "SPUE") +
  
  ylim(c(0, 1.25)) +
  
  my_theme_trans() +
  theme(base_size = 11)

SPUE
```

```{r}
day_SPUE <- sighting_day + SPUE +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(face = "bold", size = 14))
```



```{r}
ggsave(filename = "./figures/SPUE.png", SPUE, width = 1000, height = 1000, dpi = 200, units = "px")
ggsave(filename = "./figures/day_SPUE_comb.png", day_SPUE, width = 3500, height = 2000, dpi = 300, units = "px")
```




Check how far the sightings are off the observer -> count vs distance

```{r}
location <- unique(spots$spot_name)

# for (l in location) {
#   d <- data %>% 
#     filter(spot_name == l)
#   
#   plot <- d %>% 
#     ggplot()+
#     geom_histogram(aes(
#       x = distance
#     ), fill = darkblue) +
#     labs(
#       title = paste("Distance vs Number of sighting at", l)
#     ) +
#     my_theme_trans()
# 
# 
#   ggsave(filename = paste0("./figures/distance_",l, ".png"), plot, width = 1920, height = 1920, dpi = 200, units = "px")
#   
# }

data$spot_f <- factor(data$spot_name, levels = c("SandyBay", "Europa", "PrinceGeorge"))

distance <- data %>% 
  filter(spot_name %in% active_spots$spot_name) %>% 
  ggplot() + 
  geom_histogram(aes(
    x = distance,
    fill = spot_name
  ), binwidth = 550, color = "white") +
  facet_wrap(~ spot_f) +
  
  scale_fill_manual(values = c(
    "PrinceGeorge" = midblue,
    "Europa" = darkblue,
    "SandyBay" = lightblue
  )) +
  
  labs(
    x = "Distance in m",
    y = "n of Sightings",
    fill = "Spot"
  ) +
  
  my_theme_trans() +
  theme(
    strip.text  = element_text(face = "bold", colour = "black"),
    panel.border  = element_rect(fill = "transparent", colour = "black"))

distance
```

```{r}
ggsave(filename = "./figures/distance_facet.png", distance, width = 1920, height = 1020, dpi = 300, units = "px")
```


# Creating a Kernel Density Map

This map only uses the sighting data and does not correct for effort time!

```{r}

# Select on effort sightings
on <- data %>% filter(on_effort == TRUE & !is.na(sighting_poi)) %>% 
  filter(species %in% c("delphinidae", "common_dolphin")) # Filter out species that are not identified and fin whale sightings
message(nrow(on), " on-effort sightings parsed")

# Reproject so lon lat -> metres (Use same df as on otherwise error in kde)
sight_coords <- st_coordinates(on$sighting_poi)

# Weight the sightings by group size, larger group = larger weoght
weights <- on$group_size
weights[is.na(weights)] <- 1

# Transform into a factor that can be used by ks
weights <- weights * length(weights) / sum(weights)


# Fit the KDE
kde_fit <- kde(x = sight_coords, w = weights, H = Hpi(sight_coords),
               binned = FALSE, gridsize = c(100, 100))


# Turn the kde df into a tidy table
grid_x  <- kde_fit$eval.points[[1]]
grid_y  <- kde_fit$eval.points[[2]]
density <- expand.grid(x = grid_x, y = grid_y)
density$z <- as.vector(kde_fit$estimate)
density$z[density$z < max(density$z) * 0.01]

# Create df for sightings
sightings_df <- data.frame(x = sight_coords[, 1], y = sight_coords[, 2],
                           group_size = on$group_size)

# Get 95% (home range) and 50% (core area)  contour levels
hr_lev   <- contourLevels(kde_fit, cont = 95)   # 95% home range
core_lev <- contourLevels(kde_fit, cont = 50)   # 50% core area


```

### Visualise the density map

```{r}
density_map <- ggplot() +
  
  # Plot density data
  geom_raster(data = density, aes(x, y, fill = z)) +
  
  # Contour the home range and core area
  geom_contour(data = density, aes(x = x, y = y, z = z, colour = "95% home range"),
               breaks = hr_lev,   linewidth = 0.7) +
  geom_contour(data = density, aes(x = x, y = y, z = z, colour = "50% core area"),
               breaks = core_lev, linewidth = 0.9, linetype = "32") +
  
  
  # Overlay sightings
  geom_point(data = sightings_df, aes(x, y, size = group_size),
             colour = "white", alpha = 0.35) +
  
  # Add shapefile overlay
  geom_sf(data = coast, fill = "lightgray") +
  
  # Overlay spots
  geom_point(data = spots %>% 
               filter(spot_name %in% active_spots$spot_name),
             aes(x = lon, y = lat),
             shape = 25, fill = "gray10") +
  
  scale_size_continuous(range = c(1.5, 8), name = "Group size") +
  
    scale_colour_manual(
    name   = NULL,
    values = c("95% home range" = "white", "50% core area" = "white"),
    breaks = c("95% home range", "50% core area")
  ) +
  
  coord_sf(xlim = c(-5.438, -5.248), ylim = c(36.05, 36.2))  +  # Adjus t maybe later to account for all sightings, also outside the BGTW
  
  labs(
    fill = "Density",
    x = "",
    y = ""
  ) +
  # Add the scale bar
  annotation_scale(
    location = "bl",            # "bl" = bottom left
    width_hint = 0.2,          # Bar length takes up approx 40% of map width
    unit_category = "metric",   # Options: "metric", "imperial", or "nautical"
    pad_y = unit(0.05, "npc"),
    text_col  = "white"
  ) +
  
  # Add the north arrow
  annotation_north_arrow(
    location = "br", # "tr" = top right
    height = unit(0.12, "npc"),
    width = unit(0.1, "npc"),
    which_north = "true",       # Align arrow to true north pole instead of map grid
    pad_x = unit(0.05, "npc"),    # X offset from the boundary
    pad_y = unit(0.05, "npc"),    # Y offset from the boundary
    style = north_arrow_nautical(text_col = "white") # Arrow graphic style
    
  ) +
  
  my_theme_trans() +
  
  
  # Customize the colorbar, place on top and adjust spacing
  guides(
    fill = guide_colorbar(
      barwidth = unit(0.4, "npc"),
      barheight = unit(0.015, "npc"),
      title.position = "top",
      theme = theme(
        legend.ticks = element_line(color = "lightgrey", linetype = 9),
        legend.text.position = "top"
      ), order = 1
    ),
    
    size = guide_legend(override.aes = list(alpha = 1), order = 2),
    
    color = guide_legend(order = 3)
  ) +
  
  
  
  theme(legend.direction = "horizontal",
        legend.position = "inside",
        legend.position.inside = c(0.25, 0.91),
        #legend.background = element_roundrect("red", r = 0.25),
        legend.spacing = unit(0, "cm"),
        legend.margin = margin(t = 0, r = 0, b = 7, l = 0, unit = "pt"),
        text = element_text(size = 10)
    )

 windows(width = 1920/25.4,
         height = 1820/25.4)


density_map
```

```{r}
# ggsave(filename = "./figures/density_map.png", density_map, width = 1920, height = 1820, dpi = 200, units = "px")

ggsave(filename = "./figures/density_map.png", density_map, width = 3200, height = 3032, dpi = 300, units = "px")
```


Export as csv to visualize in blender
```{r}
library(sf)

# --- KDE grid -> long format (x, y, density) ---
gx <- density$eval.points[[1]]
gy <- density$eval.points[[2]]
grid <- expand.grid(x = gx, y = gy)
grid$density <- as.vector(density$estimate)

# shared local origin (note these numbers down)
ox <- min(grid$x); oy <- min(grid$y)
grid$x <- grid$x - ox
grid$y <- grid$y - oy
write.csv(grid, "kde_grid.csv", row.names = FALSE)

# --- coastline, SAME origin shift so it aligns ---
coast_utm <- st_transform(coast, 32630)        # your bay landmass polygon
co <- as.data.frame(st_coordinates(coast_utm)) # gives X, Y, L1[, L2, L3]
co$X <- co$X - ox
co$Y <- co$Y - oy
write.csv(co, "coastline.csv", row.names = FALSE)
```

```{r}
png(density, "density.png")
```


# Create detection function per spot

Clean data into clean df with only dolphin sightings and area size 

```{r, clean data for ds}
library(Distance)

# Run wedge chunk before this
sectors_m2 <- st_transform(sectors, 32630)            # back to metres
sectors_m2$area_m2 <- as.numeric(st_area(sectors_m2))

sectors_m2 <- as.data.frame(sectors_m2) %>% 
  select(-geometry) %>% 
  rename("Region.Label" = name)

sight <- data %>% 
  filter(on_effort == TRUE) %>% 
  filter(date.x > "2026-05-28") %>% 
  filter(species %in% c("common_dolphin", "delphinidae")) %>% 
  transmute(survey_id, distance, size = group_size, species)
  

eff <- effort_log %>% 
  mutate(date = dmy(date)) %>% 
  filter(date > "2026-05-28") %>% 
  mutate(sector_deg = abs(spot_bearing2 - spot_bearing1)) %>% 
  as.data.frame() %>% 
  select(-geometry)%>% 
  filter(spot_name %in% active_spots$spot_name) %>% 
  mutate(Effort = sector_deg/360) %>% 
  select("Region.Label" = spot_name,
         "Sample.Label" = survey_id,
         "height" = observer_height,
         Effort,
         wave_height, wind_dir, wind_speed, glare, sea_state) %>% 
  left_join(sectors_m2, by = "Region.Label")

dis <- eff %>% 
  as.data.frame() %>% 
  left_join(sight, by = c("Sample.Label" = "survey_id")) %>% 
  rename(Area = area_m2) %>% 
  mutate(
    size = replace_na(size, median(size, na.rm = TRUE))
  )

dis$object <- NA_integer_
dis$object[!is.na(dis$distance)] <- seq_len(sum(!is.na(dis$distance)))

saveRDS(dis, "./temp/dis.rds")

```

```{r}
tt(head(dis, n = 10), digits = 2) %>% 
  style_tt(i = 0, bold = TRUE)

as_tibble(dis) %>% 
  tt()

```



```{r, run different ds methods}


trunc <- 8000
conversion.factor <- convert_units("Metre", NULL, "square metre")

# Overall detection function (pooled)
d.hn <- ds(data=dis %>% filter(!is.na(dis$size)), key="hn", adjustment= NULL,
              transect="point", convert_units=conversion.factor, truncation=trunc)
d.hr <- ds(data=dis %>% filter(!is.na(dis$size)), key="hr", adjustment= NULL,
              transect="point", convert_units=conversion.factor, truncation=trunc)


AIC(d.hn, d.hr)

summary(d.hn)
summary(d.hr)


gof_ds(d.hn)
gof_ds(d.hr)

plot(d.hn, pdf = TRUE)
plot(d.hr, pdf = TRUE)

# height as a covariate
d.ht <- ds(data=dis %>% filter(!is.na(dis$size)), key="hr", adjustment= NULL,
              transect="point", convert_units=conversion.factor, truncation=trunc, formula = ~height)
# Region as covariate
d.re <- ds(data=dis %>% filter(!is.na(dis$size)), key="hr", adjustment= NULL,
              transect="point", convert_units=conversion.factor, truncation=trunc, formula = ~Region.Label)
#Region and height
d.rh <- ds(data=dis %>% filter(!is.na(dis$size)), key="hr", adjustment= NULL,
              transect="point", convert_units=conversion.factor, truncation=trunc, formula = ~height, ~Region.Label)

AIC(d.ht, d.re, d.rh)

# GOF to test if model fits the data
gof_ds(d.ht)
gof_ds(d.re)
gof_ds(d.rh)


plot(d.ht, pdf = TRUE)
plot(d.re, pdf = TRUE)
plot(d.rh, pdf = TRUE)

# Summary of the best model and abundance and density estimates
sum_ht <- summary(d.ht)


summarize_ds_models(d.hn, d.hr, d.ht, d.re, d.rh, output = "plain")



# Abundance estimates from detection function with stratum as stratification
dht2(d.ht, flatfile = dis, strat_formula = ~Region.Label, convert_units = conversion.factor)


```



```{r}
png(file="./figures/d_hn.png",
width=2500, height=2000, res = 300)
plot(d.hn, pdf = TRUE)
dev.off()

png(file="./figures/d_hr.png",
width=2500, height=2000, res = 300)
plot(d.hr, pdf = TRUE)
dev.off()

png(file="./figures/gof_ht_co.png",
width=2500, height=2000, res = 300)
gof_ds(d.ht)
dev.off()
```



```{r, Overlay sight bathymetry}
library(terra)
library(tidyterra)

bath <- rast("./data/GIS/BGTW_bathymetry.tif")

bath <- terra::shift(bath, dx = -455.1, dy = -230) 
bath <- terra::rescale(bath, fx = 1.21, fy = 1.28, x0 = 290271, y0 = 3998441)

#plot(bath)

# add to spat raster maxcell = 2e6

bath_map <- ggplot() +
  geom_spatraster_rgb(data = bath, maxcell = 2e6)+
  geom_sf(data = coast, fill = "lightgrey") +
  geom_sf(data = st_set_geometry(data, "geometry") %>% filter(spot_name %in% active_spots$spot_name),     shape = 25, fill = "gray10") +
  geom_sf(data = st_set_geometry(data, "sighting_poi"),
          aes(fill = species), shape = 21, colour = "white", size = 3, alpha = .6) + 
  coord_sf(crs = 4326, default_crs = 4326,
           xlim = c(-5.41, -5.28), ylim = c(36.05, 36.16))+
  
  scale_fill_manual(values = sp_fills, name = "Species",
                    labels = sp_labels) +
  
  my_theme_trans() +
  theme(
    legend.text              = ggtext::element_markdown(colour = "white"),
    legend.title = element_text(colour = "white"),
    legend.position = "inside",
    legend.position.inside = c(0.9, 0.2),
    legend.background = element_rect(fill = alpha(darkblue, alpha = 0.6))
  )

```

```{r}
ggsave(filename = "./figures/bath_map.png", bath_map, width = 3200, height = 3032, dpi = 300, units = "px")
```

