---
title: "Visualisation"
format: html
editor: source
---

```{r}
library(tidyverse)
library(readxl)
library(lubridate)
library(sf)
library(ggspatial)
library(osmdata)
library(ggforce)
library(gt)
library(ggtext)
library(legendry)
library(ks)  # Kernel smoothing for Kernel density Map
library(patchwork)
library(flextable)
library(ggfun)
library(ggsignif)
library(ggmagnify)


```

#### Colorpalette

```{r}
darkblue = "#212842"
midblue = "#4E5B9C"
lightblue = "#5C9EAD"

```

```{r}
# Excel sheet of the spots with details
spots <- read_excel(path = "./data/datasheet_landsurvey.xlsx", sheet = "spots")


# Gib Bay coastline
coast <- st_read("./data/bayofalgib.gpkg")

# Clean the sighting data and import
source("./land_survey.R")


# Set active spots as variable
active_spots <- spots %>%  
  filter(spot_name != "MedSteps" & spot_name != "WestSide") %>% 
  select(spot_name)

```

# Create a map of the survey sectors

This maps shows the sectors that were surveyed.

```{r}
make_wedge <- function(x0, y0, r, start, end, crs, n = 120) {
  b   <- seq(start, end, length.out = n) * pi / 180
  pts <- cbind(x0 + r * sin(b), y0 + r * cos(b))   # compass: x=sin, y=cos
  ring <- rbind(c(x0, y0), pts, c(x0, y0))         # close through the centre
  st_sfc(st_polygon(list(ring)), crs = crs)
}

proj <- 32630   # UTM zone 30N — metres, correct for Gibraltar

R = 6.378e+6


centres <- as.data.frame(spots)
centres <- centres %>% 
  rename(name = spot_name,
         start = spot_bearing1,
         end = spot_bearing2) %>% 
  mutate(r_m = sqrt(2*R*observer_height/20))


# project the centres into metres
xy <- st_coordinates(
  st_transform(st_as_sf(centres, coords = c("lon", "lat"), crs = 4326), proj)
)

sectors <- do.call(rbind, lapply(seq_len(nrow(centres)), function(i) {
  g <- make_wedge(xy[i, 1], xy[i, 2], centres$r_m[i],
                  centres$start[i], centres$end[i], crs = proj)
  st_sf(name = centres$name[i], geometry = g)
}))

sectors <- st_transform(sectors, 4326)   # back to lon/lat for the map

sectors <- sectors  %>% 
  filter(name != "MedSteps" & name != "WestSide")

# Sector map
sector_map <- ggplot() +
  geom_sf(data = coast) +
  geom_sf(data = sectors, aes(fill = name), colour = "black", alpha = 0.6) +
  geom_sf(data = st_set_geometry(data, "geometry") %>% filter(spot_name %in% active_spots$spot_name), shape = 25, fill = "gray10") +
  coord_sf(
    xlim = c(-5.438, -5.248),
    ylim = c(36.05, 36.195)
  ) +
  
  labs(fill = "Sectors") +
  
  # Add the scale bar
  annotation_scale(
    location = "bl",            # "bl" = bottom left
    width_hint = 0.2,          # Bar length takes up approx 40% of map width
    unit_category = "metric",   # Options: "metric", "imperial", or "nautical"
    pad_y = unit(0.05, "npc")
  ) +
  
  # Add the north arrow
  annotation_north_arrow(
    location = "br", # "tr" = top right
    height = unit(0.12, "npc"),
    width = unit(0.1, "npc"),
    which_north = "true",       # Align arrow to true north pole instead of map grid
    pad_x = unit(0.05, "npc"),    # X offset from the boundary
    pad_y = unit(0.05, "npc"),    # Y offset from the boundary
    style = north_arrow_nautical # Arrow graphic style
  ) +
  
  scale_fill_manual(values = c(
    Europa = darkblue,
    PrinceGeorge = midblue,
    SandyBay = lightblue
  )) +
  my_theme_trans()

sector_map


```

Overview map, with inset

First create inset
```{r}

europe <- map_data("world") %>% 
  filter(
    long >= -25,
    long <= 45,
    lat >= 30,
    lat <= 72
  )

inset <- ggplot(europe, aes(long, lat, group = group)) +
  geom_polygon(fill = "lightgrey", color = "black", linewidth = 0.2) +
  geom_rect(
    aes(
      xmin = -5.438,
      xmax = -5.248,
      ymin = 36.05,
      ymax = 36.195
    ),
    color = "red",
    fill = NA,
    linewidth = 1
  ) +
  annotate(
    "label",
    x = -8,
    y = 35.9,
    label = "Strait of Gibraltar",
    size = 2.5,
    fontface = "italic"
  ) +
  annotate(
    "text",
    x = -3,
    y = 38.950,
    label = "Spain",
    size = 5
  ) +
  coord_fixed(
    xlim = c(-10, 0),
    ylim = c(35, 40)
  ) +
  theme_void(base_size = 5) +
  theme(panel.border = element_rect(fill = NA, colour = "black", linewidth = 0.5),
        panel.background = element_rect(fill = "white"))
```

Insert the inset with `patchwork`

```{r}

overview <- sector_map + inset_element(inset,left = 0.00001, bottom = 0.8, right = 0.3, top = 1)

overview
```



```{r, Export}
ggsave(filename = "./figures/sector_map.png", sector_map, width = 1920, height = 1920, dpi = 200, units = "px")

ggsave(filename = "./figures/overview_map.png", overview, width = 1920, height = 1920, dpi = 200, units = "px")

```

# Visualise the sightings

::: callout-warning
Old script
:::

```{r}

# Create labels for the different spots
spot_label <- data %>% 
  select(spot_name, geometry, observer_height) %>% 
  mutate(y_nudge = case_when(
    spot_name == "SandyBay" ~ 0.003,
    spot_name == "Europa" ~ 0.003,
    spot_name == "WestSide" ~ 0.003,
    
    TRUE ~ - 0.003),
    
    x_nudge = case_when(
      spot_name == "WestSide" ~ -0.002,
      spot_name == "MedSteps" ~ 0.002,
      
      TRUE ~ 0
    )
  )


sightings <- ggplot() +
  geom_sf(data = coast, fill = "lightgray", alpha = .8) +
  geom_sf(data = st_set_geometry(data, "geometry")) +   # Set geometry so aes does not missalign
  # geom_sf(data = sectors, aes(fill = name), colour = "black", alpha = 0.4) +
  geom_sf(data = st_set_geometry(data, "sighting_poi"), #%>%  filter(date.x >= "2026-06-09"),
          aes(fill = species, linetype = species ), colour = "black", size = 3, shape = 21) +
  # geom_sf_label(data = st_set_geometry(data, "geometry"),
  #               aes(label = spot_name), size = 2, nudge_y = spot_label$y_nudge) +
  # geom_sf_text(data = st_set_geometry(data, "sighting_poi"),
  #             aes(label = species), size = 3, nudge_y = -0.003) +
  #facet_wrap(~ sector) +
  
  
  coord_sf(
    xlim = c(-5.438, -5.248),
    ylim = c(36.05, 36.195)
  ) +
  labs(fill = "Species",
       linetype = "Whales") +
  theme_minimal() +
  
  scale_shape_manual(values = c(21, 22, 24)) + 
  
  scale_fill_manual(values = c(     # If only one value is defined the rest is grayed out by default
   delphinidae = "lightblue",
   common_dolphin = "darkblue",
   striped_dolphin = "blue",
   fin_whale = "purple"
   
  ),
  labels = c(
     delphinidae = "Delphinidae",
     common_dolphin = "<b>Common Dolphin</b> <br> <i>Delphinus delphis</i>",
     striped_dolphin = "Striped Dolphin",
     fin_whale = "<b>Fin whale</b> <br> <i> Balaenoptera physalus </i>"
     
   ))+
  
  my_theme_trans() +
  
  theme(
  legend.text = ggtext::element_markdown()
)

sightings


```

### Include the boat data and combine to one df

```{r}

# Get the boat data from the data cleaning script
boat <- readRDS("./temp/boat_data.rds")

boat <- boat %>% 
  select(date.x, species, group_size, behaviour, geometry) %>% 
  mutate(
    source = "boat"
  )

land <- data %>% 
  select(date.x, species, group_size, behaviour, sighting_poi) %>% 
  rename(geometry = sighting_poi) %>% 
  mutate(
    source = "land"
  )

full <- boat %>%
  bind_rows(land)

saveRDS(full, "./temp/full.rds")

st_set_geometry(full, "geometry")


# species & their groups, in display order
sp_levels <- c("common_dolphin", "striped_dolphin", "delphinidae", "fin_whale", "unidentified")
sp_groups <- c("Odontoceti",     "Odontoceti",      "Odontoceti",  "Mysticeti", "Other")

# count sightings per species, then sum per group
sp_counts  <- table(factor(full$species, levels = sp_levels))   # per-species counts
grp_totals <- tapply(as.integer(sp_counts), sp_groups, sum)     # summed per group

# bake the totals into the group-header strings
group_headers <- paste0(sp_groups, " (n = ", grp_totals[sp_groups], ")")

species_groups <- key_group_lut(sp_levels, group_headers)


# Define shape color and legend entry once
full$species <- factor(full$species, levels = sp_levels)

# define the key contents ONCE so fill and shape stay identical
sp_labels <- c(
  common_dolphin  = "<b>Common Dolphin</b><br><i>Delphinus delphis</i>",
  striped_dolphin = "<b>Striped Dolphin</b><br><i>Stenella coeruleoalba</i>",
  delphinidae     = "<b>Delphinidae</b>",
  fin_whale       = "<b>Fin whale</b><br><i>Balaenoptera physalus</i>",
  unidentified    = "Unidentified"
)
sp_shapes <- c(common_dolphin = 21, striped_dolphin = 21, delphinidae = 21,
               fin_whale = 22, unidentified = 21)
sp_fills  <- c(common_dolphin = darkblue, striped_dolphin = midblue, delphinidae = lightblue,
               fin_whale = "purple", unidentified = "gray")


# Plot
sightings <- ggplot() +
  geom_sf(data = coast, fill = "lightgray", alpha = .8) +
  
  geom_sf(data = st_set_geometry(full, "geometry"),
          aes(fill = species, shape = species), colour = "black", size = 3) + 
  coord_sf(xlim = c(-5.50, -5.248), ylim = c(36.05, 36.195)) +
  
  scale_shape_manual(
    values = sp_shapes,
    labels = sp_labels,
    name = ""
  ) +
  
  scale_fill_manual(
    values = sp_fills,
    labels = sp_labels,
    name = ""
  ) +
  
  # Add the scale bar
  annotation_scale(
    location = "bl",            # "bl" = bottom left
    width_hint = 0.2,          # Bar length takes up approx 40% of map width
    unit_category = "metric",   # Options: "metric", "imperial", or "nautical"
    pad_y = unit(0.05, "npc")
  ) +
  
  # Add the north arrow
  annotation_north_arrow(
    location = "br", # "tr" = top right
    height = unit(0.12, "npc"),
    width = unit(0.1, "npc"),
    which_north = "true",       # Align arrow to true north pole instead of map grid
    pad_x = unit(0.05, "npc"),    # X offset from the boundary
    pad_y = unit(0.05, "npc"),    # Y offset from the boundary
    style = north_arrow_nautical # Arrow graphic style
  ) +
  
  guides(fill = guide_legend_group(key = species_groups,
                                   override.aes = list(colour = "black")),
         shape = guide_legend_group(key = species_groups,
                                   override.aes = list(colour = "black"))) +   # <- the nesting happens here
  theme_minimal() +
  my_theme_trans() +
  theme(
    legend.text              = ggtext::element_markdown(),
    legendry.legend.subtitle = element_text(face = "bold"),   # styles the group headers
    legend.position = "inside",
    legend.position.inside = c(0.15, 0.7)
  )

sightings


```



Nesting the legend to make it easier understandable

```{r}

# species & their groups, in display order
sp_levels <- c("common_dolphin", "striped_dolphin", "delphinidae", "fin_whale", "unidentified")
sp_groups <- c("Odontoceti",     "Odontoceti",      "Odontoceti",  "Mysticeti", "Other")

# count sightings per species, then sum per group
sp_counts  <- table(factor(data$species, levels = sp_levels))   # per-species counts
grp_totals <- tapply(as.integer(sp_counts), sp_groups, sum)     # summed per group

# bake the totals into the group-header strings
group_headers <- paste0(sp_groups, " (n = ", grp_totals[sp_groups], ")")

species_groups <- key_group_lut(sp_levels, group_headers)


# Define shape color and legend entry once
data$species <- factor(data$species, levels = sp_levels)

# define the key contents ONCE so fill and shape stay identical
sp_labels <- c(
  common_dolphin  = "<b>Common Dolphin</b><br><i>Delphinus delphis</i>",
  striped_dolphin = "<b>Striped Dolphin</b><br><i>Stenella coeruleoalba</i>",
  delphinidae     = "<b>Delphinidae</b>",
  fin_whale       = "<b>Fin whale</b><br><i>Balaenoptera physalus</i>",
  unidentified    = "Unidentified"
)
sp_shapes <- c(common_dolphin = 21, striped_dolphin = 21, delphinidae = 21,
               fin_whale = 22, unidentified = 21)
sp_fills  <- c(common_dolphin = darkblue, striped_dolphin = midblue, delphinidae = lightblue,
               fin_whale = "purple", unidentified = "gray")


# Plot
sightings <- ggplot() +
  geom_sf(data = coast, fill = "lightgray", alpha = .8) +
  geom_sf(data = st_set_geometry(data, "geometry") %>% filter(spot_name %in% active_spots$spot_name), shape = 25, fill = "gray10") +
  geom_sf(data = st_set_geometry(data, "sighting_poi"),
          aes(fill = species, shape = species), colour = "black", size = 3) + 
  coord_sf(xlim = c(-5.50, -5.248), ylim = c(36.05, 36.195)) +
  
  scale_shape_manual(
    values = sp_shapes,
    labels = sp_labels,
    name = ""
  ) +
  
  scale_fill_manual(
    values = sp_fills,
    labels = sp_labels,
    name = ""
  ) +
  
  # Add the scale bar
  annotation_scale(
    location = "bl",            # "bl" = bottom left
    width_hint = 0.2,          # Bar length takes up approx 40% of map width
    unit_category = "metric",   # Options: "metric", "imperial", or "nautical"
    pad_y = unit(0.05, "npc")
  ) +
  
  # Add the north arrow
  annotation_north_arrow(
    location = "br", # "tr" = top right
    height = unit(0.12, "npc"),
    width = unit(0.1, "npc"),
    which_north = "true",       # Align arrow to true north pole instead of map grid
    pad_x = unit(0.05, "npc"),    # X offset from the boundary
    pad_y = unit(0.05, "npc"),    # Y offset from the boundary
    style = north_arrow_nautical # Arrow graphic style
  ) +
  
  guides(fill = guide_legend_group(key = species_groups,
                                   override.aes = list(colour = "black")),
         shape = guide_legend_group(key = species_groups,
                                   override.aes = list(colour = "black"))) +   # <- the nesting happens here
  theme_minimal() +
  my_theme_trans() +
  theme(
    legend.text              = ggtext::element_markdown(),
    legendry.legend.subtitle = element_text(face = "bold"),   # styles the group headers
    legend.position = "inside",
    legend.position.inside = c(0.15, 0.7)
  )

sightings


```

```{r, Export}
ggsave(filename = "./figures/sightings_legend.png", sightings, width = 1920, height = 1920, dpi = 200, units = "px")
```



# Sightings analysis

```{r}
# Filter out sightings before 28.05.2026
data_an <- data %>% filter(date.x > "2026-05-28")

data_an %>% 
  filter(on_effort == TRUE) %>% 
  summarise(n_sightings = n())

data_an %>% 
  filter(species %in% c("delphinidae", "common_dolphin")) %>% 
  summarise(.by = sector, n = n())

data_an %>% 
  filter(species %in% c("delphinidae", "common_dolphin")) %>% 
  summarise(.by = species, n = n())

effort <- effort_log %>% 
  mutate(across(c(observation_start, observation_end),
                ~ hms::as_hms(.))) %>% 
  filter(!grepl("off_effort", survey_id)) %>% 
  mutate(effort_time = as.duration(observation_end - observation_start))

cat("Watch sessions:", nrow(effort),
    "| total on-effort hours:", round(sum(effort$effort_time)/60^2, 2), "\n")

er <- data_an %>%  
  filter(on_effort == 1) %>% 
  count(spot_name, name = "sightings") %>% 
  left_join(effort %>%
              group_by(spot_name) %>% 
  summarise(hours = sum(effort_time)/60^2), by = "spot_name") %>% 
  mutate(per_hour = round(sightings/hours, 2)) %>% 
  select(spot_name, sightings, hours, per_hour) %>% 
  arrange(desc(per_hour))


# Create flextable from the data
flextable(er) %>% 
   set_header_labels(
    per_hour = "Sightings per hour",
    spot_name = "Spotname",
    sightings = "# of sightings",
    hours = "on-effort Hours"
  ) %>% 
  colformat_double(
    j = "hours",
    digits = 2
  ) %>% 
  autofit() %>% 
  theme_booktabs()


cat("\nEncounter rate by station:\n"); print(er)


# On effort hours per person
per_person <- effort_log %>% 
  mutate(duration = as.numeric(
    difftime(ymd_hms(observation_end), ymd_hms(observation_start), units = "mins")
  )) %>% 
  separate_longer_delim(crew, delim = ";") %>% 
  mutate(crew = trimws(crew)) %>% 
  summarise(
    total_mins = sum(duration, na.rm = TRUE),
    total_hours = sum(duration, na.rm = TRUE)/60,
    .by = "crew"
  )

per_person

# Watches per day
as.data.frame(effort_log) %>% 
  mutate(date = dmy(date)) %>% 
  summarise(
    n = n(),
    .by = "date"
  ) %>% 
  
  ggplot(aes(x = date, y = n)) +
  geom_col(fill = darkblue) +
  scale_x_date(
    date_breaks = "2 days",
    date_labels = "%d %b"
  ) +
  my_theme_trans() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# When did we have the most sightings?
sight_time <- data %>%
  mutate(
    time_bin = paste0(
      sprintf("%02d", hour(start_time)),
      ":",
      ifelse(minute(start_time) < 30, "00", "30")
    )
  ) %>%
  count(time_bin)
 
sighting_day <- sight_time %>% 
  ggplot(aes(x = time_bin, y = n)) +
  geom_col(fill = darkblue) +
  my_theme_trans() +
  labs(
    x = "Time",
    y = "n of sightings",
    title = "Number of sightings over the day"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

sighting_day
  
```

```{r}
ggsave(filename = "./figures/sighting_day.png", sighting_day, width = 1920, height = 1520, dpi = 200, units = "px")
```

### Does tide have an influence? 

```{r}
tide_df <- data %>% 
  mutate(tide = as.factor(tide)) %>% 
  summarise(
    n = n(),
    .by = "tide"
  ) %>% 
  ggplot(aes(x = tide, y = n)) +
  geom_col()

tide_df

```


### Compare current to target watches/hours
Target hours comes from the Power Analysis script

```{r}
per_spot <- as.data.frame(effort_log) %>% 
  filter(spot_name %in% active_spots$spot_name) %>% 
  mutate(duration = as.numeric(
    difftime(ymd_hms(observation_end), ymd_hms(observation_start), units = "mins")
  )) %>% 
  summarise(
    total_mins = sum(duration, na.rm = TRUE),
    total_hours = sum(duration, na.rm = TRUE)/60,
    percent = round((total_hours / 10) * 100, 2), 
    .by = "spot_name"
  )

aim <- data.frame(
  spot = per_spot$spot_name,
  target = 10
)

hour_spot <- per_spot %>% 
  ggplot(aes(x = spot_name, y = per_spot$total_hours)) +
  geom_col(fill = darkblue) +
  geom_text(aes(x = spot_name, y = total_hours + 0.5), ,label = paste(per_spot$percent, "%")) +
  geom_errorbar(
    data = aim,
    aes(x = spot, ymin = target, ymax = target),
    linewidth = 1.5,
    linetype = "dashed",
    color = "darkred"
  ) +
  labs(
    x = "",
    y = "Hours per spot"
  ) +
  my_theme_trans()
  
hour_spot
```

```{r}
ggsave(filename = "./figures/hour_spot.png", hour_spot, width = 1920, height = 1520, dpi = 200, units = "px")
```


### Convert to sightings per unit effort (SPUE)

```{r}
effort <- effort_log %>%
  filter(!str_detect(survey_id, "off_effort")) %>%
  mutate(
    duration_min = as.numeric(difftime(observation_end, observation_start,
                                        units = "mins"))
  )
 
# ---- 3. Detections: keep on-effort sightings -------------------------------
sightings <- detection %>%
  filter(on_effort == 1)

 
# ---- 4. Sightings & individuals per scan -----------------------------------
per_scan <- sightings %>%
  group_by(survey_id) %>%
  summarise(
    n_sightings   = n(),
    n_individuals = sum(group_size, na.rm = TRUE),  # one NA group size -> 0
    .groups = "drop"
  )
 
# Join back onto the full effort table; scans with no sighting become 0 (NOT NA)
effort <- effort %>%
  left_join(per_scan, by = "survey_id") %>%
  mutate(
    n_sightings   = replace_na(n_sightings, 0),
    n_individuals = replace_na(n_individuals, 0)
  )
 
# ---- 5. SPUE by spot -------------------------------------------------------
spue_by_spot <- effort %>%
  group_by(spot_name) %>%
  summarise(
    scans        = n(),
    effort_hours = sum(duration_min, na.rm = TRUE) / 60,
    sightings    = sum(n_sightings),
    individuals  = sum(n_individuals),
    .groups = "drop"
  ) %>%
  mutate(
    spue_per_scan   = sightings / scans,        # headline SPUE
    sightings_per_h = sightings / effort_hours, # encounter rate
    indiv_per_h     = individuals / effort_hours
  ) %>%
  arrange(desc(scans))
 
# ---- 6. SPUE overall -------------------------------------------------------
spue_overall <- effort %>%
  summarise(
    spot_name    = "TOTAL",
    scans        = n(),
    effort_hours = sum(duration_min, na.rm = TRUE) / 60,
    sightings    = sum(n_sightings),
    individuals  = sum(n_individuals)
  ) %>%
  mutate(
    spue_per_scan   = sightings / scans,
    sightings_per_h = sightings / effort_hours,
    indiv_per_h     = individuals / effort_hours
  )
 
spue_summary <- bind_rows(spue_by_spot, spue_overall)
 
# ---- 7. Report & export ----------------------------------------------------
spue_report <- spue_summary %>%
  mutate(across(c(effort_hours, spue_per_scan, sightings_per_h, indiv_per_h),
                ~ round(.x, 3)))
 
print(spue_report, n = Inf)

spue_report %>% 
  filter(spot_name %in% active_spots$spot_name | spot_name == "TOTAL") %>% 
  select(!geometry) %>% 
  flextable() %>% 
  delete_columns(j = "geometry") %>% 
  set_header_labels(
    spot_name = "Spot Names",
    scans = "Scans",
    effort_hours = "Hours",
    sightings = "Sightings",
    individuals = "Individuals",
    spue_per_scan = "SPUE (per scan)",
    sightings_per_h = "Sightings per h",
    indiv_per_h = "Indiv. per h"
  ) %>% 
  colformat_double(
    j = c("effort_hours", "sightings_per_h", "sightings_per_h"),
    digits = 2
  ) %>% 
  bold(i = 4) %>% 
  autofit() %>% 
  theme_booktabs()

```

Analize spue and compare. Test if there is significant difference between the spots (Kruskal-Wallis) and between which spots (dun-Test). Additional glm 

```{r}
library(FSA)

active <- c("Europa", "SandyBay", "PrinceGeorge")
scans  <- effort %>% filter(spot_name %in% active)   # per-scan rows, NOT the summary

count(scans, spot_name)     


kruskal.test(n_sightings ~ spot_name, data = scans)

dunnTest(n_sightings ~ factor(spot_name), data = scans, method = "bh")
```

```{r}


spue_clean <-spue_report %>% 
  filter(spot_name %in% active_spots$spot_name) %>% 
  select(!geometry)

SPUE <- spue_clean %>% 
  ggplot(aes(x = spot_name, y = spue_per_scan))+
  geom_col(fill = darkblue) +
  geom_signif(
    # Specify which columns to connect (using group names or numeric X-indices)
    comparisons = list(c("Europa", "PrinceGeorge")),
    
    # Enter your exact pre-calculated p-value symbols in the same order
    annotations = c("***"), 
    map_signif_level = TRUE,
   
    
    # Visual styling details
    tip_length = 0.02,
    textsize = 6
  ) +
  
  labs(x = "", y = "SPUE") +
  
  ylim(c(0, 1.2)) +
  
  my_theme_trans() +
  theme(base_size = 11)

SPUE
```

```{r}
ggsave(filename = "./figures/SPUE.png", SPUE, width = 1000, height = 1000, dpi = 200, units = "px")
```




Check how far the sightings are off the observer -> count vs distance

```{r}
location <- unique(spots$spot_name)

for (l in location) {
  d <- data %>% 
    filter(spot_name == l)
  
  plot <- d %>% 
    ggplot()+
    geom_histogram(aes(
      x = distance
    ), fill = darkblue) +
    labs(
      title = paste("Distance vs Number of sighting at", l)
    ) +
    my_theme_trans()


  ggsave(filename = paste0("./figures/distance_",l, ".png"), plot, width = 1920, height = 1920, dpi = 200, units = "px")
  
}



distance <- data %>% 
  ggplot() + 
  geom_histogram(aes(
    x = distance,
    fill = sector
  ), binwidth = 550, color = "white") +
  facet_wrap(~ sector) +
  
  scale_fill_manual(values = c(
    S = darkblue,
    W = midblue,
    E = lightblue
  )) +
  
  my_theme_trans() +
  theme(
    strip.text  = element_text(face = "bold"),
    panel.border  = element_rect(fill = "transparent", colour = "black"))

distance
```

```{r}
ggsave(filename = "./figures/distance_facet.png", distance, width = 1920, height = 1520, dpi = 200, units = "px")
```


# Creating a Kernel Density Map

This map only uses the sighting data and does not correct for effort time!

```{r}

# Select on effort sightings
on <- data %>% filter(on_effort == TRUE & !is.na(sighting_poi)) %>% 
  filter(species %in% c("delphinidae", "common_dolphin")) # Filter out species that are not identified and fin whale sightings
message(nrow(on), " on-effort sightings parsed")

# Reproject so lon lat -> metres (Use same df as on otherwise error in kde)
sight_coords <- st_coordinates(on$sighting_poi)

# Weight the sightings by group size, larger group = larger weoght
weights <- on$group_size
weights[is.na(weights)] <- 1

# Transform into a factor that can be used by ks
weights <- weights * length(weights) / sum(weights)


# Fit the KDE
kde_fit <- kde(x = sight_coords, w = weights, H = Hpi(sight_coords),
               binned = FALSE, gridsize = c(100, 100))


# Turn the kde df into a tidy table
grid_x  <- kde_fit$eval.points[[1]]
grid_y  <- kde_fit$eval.points[[2]]
density <- expand.grid(x = grid_x, y = grid_y)
density$z <- as.vector(kde_fit$estimate)
density$z[density$z < max(density$z) * 0.01]

# Create df for sightings
sightings_df <- data.frame(x = sight_coords[, 1], y = sight_coords[, 2],
                           group_size = on$group_size)

# Get 95% (home range) and 50% (core area)  contour levels
hr_lev   <- contourLevels(kde_fit, cont = 95)   # 95% home range
core_lev <- contourLevels(kde_fit, cont = 50)   # 50% core area


```

### Visualise the density map

```{r}
density_map <- ggplot() +
  
  # Plot density data
  geom_raster(data = density, aes(x, y, fill = z)) +
  
  # Contour the home range and core area
  geom_contour(data = density, aes(x = x, y = y, z = z, colour = "95% home range"),
               breaks = hr_lev,   linewidth = 0.7) +
  geom_contour(data = density, aes(x = x, y = y, z = z, colour = "50% core area"),
               breaks = core_lev, linewidth = 0.9, linetype = "32") +
  
  
  # Overlay sightings
  geom_point(data = sightings_df, aes(x, y, size = group_size),
             colour = "white", alpha = 0.35) +
  
  # Add shapefile overlay
  geom_sf(data = coast, fill = "lightgray") +
  
  # Overlay spots
  geom_point(data = spots %>% 
               filter(spot_name %in% active_spots$spot_name),
             aes(x = lon, y = lat),
             shape = 25, fill = "gray10") +
  
  scale_size_continuous(range = c(1.5, 8), name = "Group size") +
  
    scale_colour_manual(
    name   = NULL,
    values = c("95% home range" = "white", "50% core area" = "white"),
    breaks = c("95% home range", "50% core area")
  ) +
  
  coord_sf(xlim = c(-5.438, -5.248), ylim = c(36.05, 36.2))  +  # Adjus t maybe later to account for all sightings, also outside the BGTW
  
  labs(
    fill = "Density",
    x = "",
    y = ""
  ) +
  # Add the scale bar
  annotation_scale(
    location = "bl",            # "bl" = bottom left
    width_hint = 0.2,          # Bar length takes up approx 40% of map width
    unit_category = "metric",   # Options: "metric", "imperial", or "nautical"
    pad_y = unit(0.05, "npc"),
    text_col  = "white"
  ) +
  
  # Add the north arrow
  annotation_north_arrow(
    location = "br", # "tr" = top right
    height = unit(0.12, "npc"),
    width = unit(0.1, "npc"),
    which_north = "true",       # Align arrow to true north pole instead of map grid
    pad_x = unit(0.05, "npc"),    # X offset from the boundary
    pad_y = unit(0.05, "npc"),    # Y offset from the boundary
    style = north_arrow_nautical(text_col = "white") # Arrow graphic style
    
  ) +
  
  my_theme_trans() +
  
  
  # Customize the colorbar, place on top and adjust spacing
  guides(
    fill = guide_colorbar(
      barwidth = unit(0.4, "npc"),
      barheight = unit(0.015, "npc"),
      title.position = "top",
      theme = theme(
        legend.ticks = element_line(color = "lightgrey", linetype = 9),
        legend.text.position = "top"
      ), order = 1
    ),
    
    size = guide_legend(override.aes = list(alpha = 1), order = 2),
    
    color = guide_legend(order = 3)
  ) +
  
  
  
  theme(legend.direction = "horizontal",
        legend.position = "inside",
        legend.position.inside = c(0.25, 0.91),
        #legend.background = element_roundrect("red", r = 0.25),
        legend.spacing = unit(0, "cm"),
        legend.margin = margin(t = 0, r = 0, b = 7, l = 0, unit = "pt"),
        text = element_text(size = 10)
    )

# windows(width = 1920/25.4,
#         height = 1820/25.4)


density_map
```

```{r}
ggsave(filename = "./figures/density_map.png", density_map, width = 1920, height = 1820, dpi = 200, units = "px")
```


Export as csv to visualize in blender
```{r}
library(sf)

# --- KDE grid -> long format (x, y, density) ---
gx <- density$eval.points[[1]]
gy <- density$eval.points[[2]]
grid <- expand.grid(x = gx, y = gy)
grid$density <- as.vector(density$estimate)

# shared local origin (note these numbers down)
ox <- min(grid$x); oy <- min(grid$y)
grid$x <- grid$x - ox
grid$y <- grid$y - oy
write.csv(grid, "kde_grid.csv", row.names = FALSE)

# --- coastline, SAME origin shift so it aligns ---
coast_utm <- st_transform(coast, 32630)        # your bay landmass polygon
co <- as.data.frame(st_coordinates(coast_utm)) # gives X, Y, L1[, L2, L3]
co$X <- co$X - ox
co$Y <- co$Y - oy
write.csv(co, "coastline.csv", row.names = FALSE)
```

```{r}
png(density, "density.png")
```


# Create detection function per spot

```{r}
library(Distance)

dat <- data %>% 
  select(-observer)

# on-effort detections with a valid radial distance
d <- subset(dat, on_effort == TRUE & !is.na(distance))
d$size      <- d$group_size
d$spot_name <- factor(d$spot_name)

# truncate the furthest ~10% (reticle angular resolution is poor at km scale)
trunc <- as.numeric(quantile(d$distance, 0.90, na.rm = TRUE))

# pooled POINT-transect detection functions
m_hn   <- ds(d, transect = "point", key = "hn", adjustment = "cos", truncation = trunc)
m_hr   <- ds(d, transect = "point", key = "hr", adjustment = "cos", truncation = trunc)

# detectability as a covariate instead of per-spot fits (MCDS, no adjustment terms)
m_ht   <- ds(d, transect = "point", key = "hn", formula = ~observer_height, truncation = trunc)
m_spot <- ds(d, transect = "point", key = "hn", formula = ~spot_name,       truncation = trunc)

AIC(m_hn, m_hr, m_ht, m_spot)   # model selection
summary(m_ht)
gof_ds(m_ht)                    # Cramér–von Mises goodness-of-fit
plot(m_ht)      

```

